Characterization and modulation of human insulin degrading enzyme conformational dynamics to control enzyme activity

Version 1

Abstract

Abstract

Insulin degrading enzyme (IDE) is a dimeric 110 kDa M16A zinc metalloprotease that degrades amyloidogenic peptides diverse in shape and sequence, including insulin, amylin, and amyloid-β, to prevent toxic amyloid fibril formation. IDE has a hollow catalytic chamber formed by four homologous subdomains organized into two ∼55 kDa N- and C-domains (IDE-N and IDE-C, respectively), in which peptides bind, unfold, and are repositioned for proteolysis. IDE is known to transition between a closed state, poised for catalysis, and an open state, able to release cleavage products and bind new substrate. Here, we present five cryoEM structures of the IDE dimer at 3.0-4.1 Å resolution, obtained in the presence of a sub-saturating concentration of insulin. Analysis of the heterogeneity within the particle populations comprising these structures combined with all-atom molecular dynamics (MD) simulations permitted a comprehensive characterization of IDE conformational dynamics. Our analysis identified the structural basis and key residues for these dynamics that were not revealed by IDE static structures. Notably arginine-668 serves as a molecular latch mediating the open-close transition and facilitates key protein motions through charge-swapping interactions at the IDE-N/C interface. Our size-exclusion chromatography-coupled small-angle X-ray scattering and enzymatic assays of an arginine-668 to alanine mutant indicate a profound alteration of conformational dynamics and catalytic activity. Taken together, this work highlights the power of integrating experimental and computational methodologies to understand protein dynamics, offers the molecular basis of unfoldase activity of IDE, and provides a new path forward towards the development of substrate-specific modulators of IDE activity.

Introduction

Protein homeostasis (a.k.a. proteostasis) is maintained by three primary mechanisms: chaperones, ubiquitination/proteasome, and autophagy (1, 2). Disruptions in proteostasis can lead to the accumulation of amyloid fibrils and subsequent human diseases (3–5). As a result, many proteases have evolved to specifically target amyloidogenic peptides (6, 7). Among those, the cryptidase family, which includes the M16 metalloproteases insulin degrading enzyme (IDE) use an internal catalytic chamber, or “crypt”, to capture and selectively degrade the monomeric form of amyloid peptides to control the formation of amyloid fibrils (6, 8). IDE effectively degrades various bioactive peptides, including amyloid-β (Aβ), a peptide associated with the progress of Alzheimer’s disease, and three blood glucose-regulating hormones: insulin, amylin, and glucagon (9). Consequently, defects in IDE alter the progression of type 2 diabetes mellitus and Alzheimer’s disease in animal models and are linked to these diseases in humans (9–17). IDE is a promising therapeutic target, as its inhibition improves glucose tolerance, yet progress has been hampered by the contrary actions of its diverse substrate pool (9, 18, 19). For example, IDE overexpression has been shown to reduce Aβ loads, in mice, however, this reduction is accompanied by hyperglycemia from a similar reduction in insulin levels (20). Therefore, a deeper understanding of the molecular mechanisms underlying IDE catalysis is required to facilitate the development of future therapeutics.

Integrative structural approaches have provided key insights into how IDE selectively degrades amyloidogenic peptides (6, 9, 21). IDE is a dimeric 110 kDa metalloprotease. Each protomer is comprised of four homologous subdomains organized into two ∼55 kDa domains, IDE-N and IDE-C, connected by a short flexible linker (22, 23). Our previous cryoEM analysis revealed the major conformational states of IDE. In the absence of substrate, each protomer was found to adopt an open (O) or partial open (pO) state which are primarily differentiated by the displacement of IDE-N relative to IDE-C, the combination of which led to three dimeric conformations (O/O, O/pO, pO/pO) (21). It was observed that substrate binding induced both protomers of IDE to fully close, referred to as the partial closed (pC, pC/pC for the dimer) state, as the cryoEM closed state was found to be slightly more open than previously solved crystal structures, likely influenced by the constraints of the crystal lattice (22, 24). Only the O state permits substrate access to the catalytic chamber, yet the catalytic cleft is stabilized only in the pC state with substrate bound, requiring IDE to undergo a substantial open-close transition during catalysis, the specifics of which are unknown (21).

It has been established that many IDE substrates must be unfolded prior to cleavage but, as IDE substrates are highly diverse in size and structure, the mechanistic basis of this unfoldase activity has proven elusive. We hypothesized that the unfoldase activity of IDE is mediated by the relative motions of IDE-N and IDE-C. To test this hypothesis, we solved cryoEM structures of IDE with a 2:1 IDE:insulin ratio and used recently developed computational approaches to deconstruct the particle heterogeneity within our structures to infer conformational dynamics. Integration of this experimental data with all-atom molecular dynamics (MD) simulations allowed us to identify and modulate the molecular interactions governing key conformational dynamics to manipulate the enzymatic activity of IDE in vitro.

Results

CryoEM structures of IDE in the presence of a sub-saturating concentration of insulin

We have previously used cryoEM to solve the structure of IDE in the presence of a 5-fold molar excess of insulin, resulting in a dominant pC/pC state with both substrate binding chambers yielding density corresponding to unfolded insulin (21). To explore the mechanism of IDE-directed substrate unfolding under more physiologically relevant conditions, here we investigated IDE-insulin interactions with an IDE:insulin molar ratio of 2:1. We collected a dataset of ∼7,600 micrographs on a Titan Krios operated at 300 keV, from which 7.2 million particles were picked for processing in RELION. Following an established workflow, we generated 5 structures: three previously observed structures with improved resolution (pC/pC, O/O, and O/pO at 3.0 Å, 3.8 Å, and 4.1 Å resolution) and two novel states (O/pC and pO/pC at 3.4 Å and 3.3 Å resolution, respectively) (Fig. 1A, S1, Table 1). Interestingly, the pC/pC was still found to be dominant, despite IDE and insulin being present at a 2:1 molar ratio, consistent with the allostery of IDE (21,24).

CryoEM structures.

(A) Overview of the cryoEM structures. See figure S1 for processing details. (B) Comparison of the open (O), partial open (pO), and partial closed (pC) subunit states present in our cryoEM structures with domain organization. The distance between the D1 and D4 domain centers-of-mass (D1-D4 COM) along with the dihedral angle formed by the D1-D2-D3-D4 domain centers-of-mass (D1-D2-D3-D4 dihedral) described in Zhang et al. (21) and depicted in Fig. S4 were used as biologically important criteria to quantify observed conformations. (C) Insulin density and corresponding model in our cryoEM structures. Both the A chain (magenta) and B chain (yellow) can fit the density in the exosite and catalytic cleft.

The individual protomers of IDE adopt the same three conformations, O, pO, and pC states reported previously but there are significant differences in the O and pO states (Fig. 1B) (21). While the pC/pC state is nearly identical to the two states reported previously (RMSD of ∼0.6 Å; PDB ID: 6BFC, 6B3Q) (21), the O and pO states appear more closed when insulin is present, suggesting that the presence of insulin promotes the open to closed transition. Specifically, the pO and O subunits are 3-5 Å more closed based on the distance between the centers-of-mass (COM) of the IDE D1 and D4 subdomains, while the dihedral angle formed by the COM of the D1-D2-D3-D4 subdomains is reduced by 5-10° (Fig. 1B). Alignment of the new O/O and O/pO states to their previously solved counterparts reveals little difference, with a global RMSD of 1.316 Å (O/O, PDB ID: 6B7Y) and 1.405 Å (O/pO, PDB ID: 6BF8), respectively (21). The primary source of variation between the different states remains the degree of opening between the two domains of each subunit, as a global alignment of IDE-N and IDE-C across all structures reveal the domains primarily act as rigid bodies (RMSD <1.4 Å). Consistent with previous observations, the door region, which contains the catalytic zinc binding site, exhibited higher B-factors in the O states than in either the pO or pC state (21) (Fig. S2).

All five cryoEM structures presented here contain four subunits that have clear density present at the catalytic cleft and exosite indicative of bound substrate, thus adopting the pC state. Previous structures have modeled the same chain of insulin in both the exosite and catalytic cleft, but the best models we could build into our density suggest that the chain bound to the exosite is not the same chain positioned for cleavage at the catalytic cleft (Fig. 1C). The preponderance of data suggests it to be unlikely that insulin adopts a preferred orientation within the catalytic chamber, rather the catalytic chamber can accommodate insulin in four different orientations, with either the A or B chain bound to the exosite and catalytic cleft in cis and trans orientations. This promiscuity of binding has been theorized as a reason for why there is no chain preference for the initial insulin cleavage event (25). Crystal structures of insulin are highly compact and globular, with little spatial separation between the N- and C-termini of each chain, and the IDE cleavage sites are located within α-helices (26). In its crystallized state, insulin cannot interact with both the exosite and catalytic cleft unless it unfolds, at least partially. Recent work used MD simulations to study the conformational dynamics of insulin in solution and identified several major “elements of disorder” to describe the partially unfolded intermediate structures they observed (27). We docked these structures of insulin into the closed catalytic chamber of IDE and found that the chamber was able to easily accommodate the positioning of either chain near the exosite and catalytic cleft in cis and trans orientations (Fig. S3).

Structural heterogeneity of cryoEM structures

Numerous structures of IDE exist with subunits adopting either an open or closed conformation, yet we lack an understanding of how IDE transitions from an open to a closed state. The simplest explanation of this transition, based on structural data, would be a direct, rigid body translation of IDE-N relative to IDE-C, and this is the predominant model within the field (21, 22). To better understand what this transition would look like, we measured and plotted the changes in the D1-D4 COM distance versus the changes in the D1-D2-D3-D4 COM dihedral angle for all available cryoEM structures of IDE (Fig. S4). We found that in the absence of substrate, the transition pathway produced a linear relationship between states. However, when our structures generated in the presence of insulin were analyzed, the linear relationship no longer held (Fig. S4), suggesting that the simple linear transition typically presumed for the open-close transition between distinct IDE states does not accurately depict the complexity of IDE dynamics.

Recently, several approaches have been developed to understand the conformational heterogeneity present within cryoEM data. We employed multibody analysis in RELION and 3D variability analysis (3DVA) in cryoSPARC to investigate the conformational heterogeneity within our particle populations (28, 29). Of these, the range of motion predicted by multibody analysis most closely matched the reported structures while 3DVA exhibited consistent results across a smaller magnitude of transition (Fig. S4-S6). Therefore, we focused on multibody analysis below. Multibody analysis models the structural heterogeneity in our data as the result of motions of independent, user-defined rigid bodies. As discussed above, the primary source of structural variation in our structures is the positioning of IDE-N relative to IDE-C. We see essentially no structural changes within the domains, suggesting the domains behave as rigid bodies, an assertion further supported by our all-atom molecular dynamics (MD) simulations (Fig. 2A). Multibody analysis defines the principal components of variance along discrete eigenvectors, which we interpret as proxies for the dominant components of molecular motion. We quantified the change in D1-D4 COM distance and D1-D2-D3-D4 COM dihedral of representative structures along the heterogeneity gradients of the top 9 eigenvectors for each of our IDE structures, representing ∼70-90% of the total structural heterogeneity per structure. The results indicated that the particles comprising our cryoEM structures sampled a significantly greater conformational space than would be expected from analysis of the ensemble structures alone (Fig. S5). Most intriguingly, the conformational changes displayed an unexpectedly high degree of change in the D1-D2-D3-D4 COM dihedral angle, indicating a significant rotation of IDE-N relative to IDE-C (Fig. 2B,C, Fig. S5). Integration of the multibody results across all states of IDE suggest two dominant components of structural variance. First is a translation-dominant conformational change wherein IDE-N swings toward or away from IDE-C as if mediated by a hinge formed by the interdomain linker region of IDE (Fig. 2D, Movie S1). This conformational change closely resembles our understanding of the IDE open-closed transition inferred from analysis of the ensemble structures. The second component is a rotation-dominant conformational change wherein IDE-N rotates orthogonal to the plane of the dimer as if it were being screwed into or ground against IDE-C (Fig. 2E, Movie S2). Importantly, these dominant components of structural variance were also observed when the particle populations were analyzed with cryoSPARC’s 3DVA implementation, albeit over a smaller magnitude of conformational change (Fig. S6). These dominant components also correlate well with the lowest frequency modes revealed by normal mode analysis (Movies S3, S4). Interestingly, while the presence of insulin was found to substantially influence the consensus reconstructions of IDE, we found that the presence or absence of insulin had no significant effect on the principal components of structural variance for IDE, consistent with the current understanding that enzyme conformational changes are “hardwired” into the structure (30, 31).

Conformational dynamics of IDE implied by structural heterogeneity.

(A) All-atom MD simulations analysis. The primary source of structural variance (RMSD) results from the IDE-N moving against IDE-C as a rigid body. Rigid bodies were defined as colored for multibody refinement in RELION. (B-C) Multibody analysis. The range of conformational variance described by the top principal component vectors displays an unexpectedly high degree of rotational motion, as measured by the change in D1-D2-D3-D4 dihedral angle across each vector’s gradient of structural heterogeneity, in both the absence (B) and presence (C) of insulin compared to the expected open-close transition pathway predicted from a linear interpolation of the experimentally determined structures of IDE (dashed line, Fig. S4). Two dominant components of structural variance are revealed from multibody analysis: (D) where IDE-N swings relative to IDE-N about the inter-domain linker, and (E) where IDE-N rotates against IDE-C. Starting (orange) and ending (red) states of IDE-N shown with pathway depicted by arrows. IDE-C shown as gray surface.

Molecular basis of the open-close transition revealed by all-atom MD simulations

Analysis of the particle heterogeneity comprising cryoEM structures provides excellent information about the conformational space sampled by the protein of interest. However, such analysis lacks a temporal component, and thus only implies motions. To overcome this barrier and examine how the conformation of IDE changes over time, we performed six replicate all-atom MD simulations with explicit solvent. Simulations were started from the O/pO state (PDB: 7RZG), with the missing loops modeled in to generate continuous peptide chains and run for one microsecond. By starting with the O/pO structure, we investigated the dynamics of both the open and closed subunits simultaneously. We observed that the conformational space sampled by our simulations correlated well with the conformational space sampled by our cryoEM particle population, as revealed by multibody analysis (Fig. S7). In 5 of our 6 simulations, the open subunit closed quickly, typically in <200 ns (Fig. 3A). Analysis of the open-close transition revealed two key findings. First, the open subunits did not close to a singular structure. We define closing as a subunit reaching a similar D1-D4 COM distance as the pO or pC states, but we observed distinct differences among the subunits indicative of IDE-N rotation relative to IDE-C. These differences are in line with our second key finding, that the conformational changes associated with the open-close transition did not follow the same general pathway among the simulations, as we observed a high degree of variability upon closing (Fig. 3B). Despite this variability, R668 consistently stood out as one of the first IDE-C residues to form new interactions with IDE-N, either D309 or E381 (Fig. 3C). In most simulations, formation of this interaction preceded a large decrease in the D1-D4 COM distance (Fig. 3D, Fig. S8). Interestingly, these R668 interactions are not present in the ensemble cryoEM or crystal structures of IDE. We hypothesize that the biochemical properties and spatial localization of R668 enable it to essentially reach out and grab onto the N-domain, initiating the formation of a complex hydrogen bonding network that drives the open-close transition towards completion. Previously published hydrogen-deuterium exchange mass spectrometry (HDX-MS) data indicates that R668 is stabilized under conditions that promote IDE closing, suggesting that R668 may function as a sort of molecular latch (Fig. 3E)(21).

All-atom MD reveals a molecular basis for IDE conformational dynamics.

(A) Measurements of the O subunit D1-D4 distance over the course of six separate microsecond long all-atom MD simulations of WT IDE. Of which, the open subunit closed in 5 of the 6 simulations. (B) Plot of the O subunit D1-D4 distance vs the D1-D2-D3-D4 COM dihedral angle over the course of the simulation of WT IDE. The open subunits displayed a variety of closing pathways and did not close to a consensus structure. Starting structure shown as black dot, pO structure shown as white dot. (C) R668 acts as a guidepost residue, rapidly interacting with D309 or E381. Formation of this interaction is associated with rapid closing, as measured by a decrease in D1-D4 distance (D). (E) Hydrogen-deuterium exchange mass spectrometry highlights the importance of R668 in mediating the open-close transition. In the presence of insulin (panel 1, red), Aβ (panel 2, red), and BDM-44768 (panel 3, red), all of which promote IDE closing, the peptide containing R668 shows reduced deuterium exchange relative to apo-IDE (black), yet in the presence of 6bk (panel 4, red), which does not promote closing, there is no difference in the exchange rates for the R668 containing peptide relative to apo-IDE (black). Helix containing R668 colored by red (increase) – white (no change) – blue (decrease) gradient depicting the degree of deuterium exchange relative to apo-IDE. (F) Measurements of the O subunit D1-D4 distance over the course of six separate microsecond long all-atom MD simulations of IDE R668A. (G) Plot of the O subunit D1-D4 distance vs the D1-D2-D3-D4 COM dihedral angle over the course of the simulation of IDE R668A. The six separate microsecond long simulations indicate that an R668A mutation significantly alters the closing dynamics of IDE (F) and increases the rotational motion (G) relative to WT (panels A and B respectively). Starting structure shown as black dot, pO structure shown as white dot.

To further probe the importance of R668, we ran 6 additional simulations with IDE carrying a R668 to alanine (R668A) point mutation under identical conditions as described previously. These simulations displayed significantly altered dynamics when compared to the wild-type simulations. Notably, the R668A mutant was found to close more slowly, and many simulations did not reach the D1-D4 COM distance typified by a closed state structure, instead stabilizing with a significantly larger D1-D4 COM distance than we had previously observed in our wild type (WT) simulations or our experimental structures (Fig. 3F). We also observed that the R668A mutant preferentially sampled a subset of conformational space rarely explored in our WT simulations and displayed significantly greater variation in the D1-D2-D3-D4 dihedral angle (Fig. 3G). As a result of this increased rotational exploration, several of our R668A simulations closed to a conformation where IDE-N is offset and rotated relative to IDE-C when compared to the experimentally determined structures (Fig. S9). This altered conformation results in a structure where IDE-N and IDE-C are fully engaged, but the offset produces a deceivingly high D1-D4 COM distance. With this in mind, we sought to better understand the conformational dynamics of IDE in the closed (pO) state.

R668A mediates IDE conformational dynamics in vitro

The R668A mutation was found to significantly alter the conformational dynamics of IDE’s open-close transition in our simulations (Fig. 3F, G), so we expressed and purified the R668A mutant to determine if the changes predicted from our MD simulations resulted in altered enzymatic activity relative to WT. Previous work has established that IDE adopts a dominant O/pO state in solution (21). Consistent with our MD simulations, size-exclusion chromatography revealed that the R668A construct eluted slightly earlier than WT IDE, suggesting that the R668A mutation induces a greater proportion of molecules to adopt a more open conformation, possibly shifting to a dominant O/O state (Fig. 4A). Next, we compared the enzymatic activity of the R668A and WT constructs using the fluorogenic substrate (7-methoxycourmarin-4-yl)acetyl-RPPGFSAFK(2,4-dinitrophenyl)-OH (substrate V), a bradykinin mimetic which has previously been used to characterize the enzymatic activity of IDE and related enzymes (24, 32). We found that WT IDE has activity consistent with previously published data (24) and the R668A mutation produced a ∼5-fold decrease in activity (Fig. 4B). ATP is known to promote IDE degradation of small substrates, i.e. bradykinin, but not larger substrates, i.e. insulin (33). Consistent with this observation, the addition of ATP increased the activity of both WT and R668A constructs (Fig. 4B). We also tested if the R668A mutation would alter the degradation of larger, well-folded substrates, using insulin in a direct competition assay. We found that, after accounting for the previously observed 5x decrease in activity, insulin yielded an apparent Ki of ∼8 nM for WT IDE, but this value was reduced to ∼52 nM for the R668A construct (Fig. 4C). We then performed size exclusion chromatography-coupled small-angle X-ray scattering (SEC-SAXS) for both constructs to ascertain if the observed differences in enzymatic activity could be explained by altered biophysical properties. The WT scattering profile produced an Rg of 49.4 +/- 0.9 Å, consistent with previous SEC-SAXS results (21)(Fig. 4D, E). The R668A scattering profile produced an Rg of 54.7 +/- 0.4 Å, consistent with our SEC data that R668A mutant has slightly larger hydrodynamic radius (Fig. 4D). Together, our data indicates that R668A mutation can profoundly affect the conformational dynamics governing the open-closed transition, which leads to the altered substrate binding and catalysis.

R668A alters IDE activity in vitro

(A) Elution profile of WT IDE (blue) compared to the R668A mutant (orange) from a S200 SEC column. (B) Degradation of the fluorescent substrate MCA-RPPGFSAFK(Dnp) by WT IDE and the R668A construct in the presence and absence of ATP. Data represents the average initial velocities of three replicates performed at a protein concentration of 3.125 nM. Error bars (gray) represent the standard error. (C) Inhibition of MCA-RPPGFSAFK(Dnp) degradation by WT IDE (circles, solid fit lines) and IDE R668A (squares, dashed fit lines) in the presence of varying amounts of insulin. Data was fit to the Michaelis-Menten (black) and Hill equations (red). Relevant parameters, Michaelis-Menten: WT: χ2=0.001, Vmax=0.951, Ki=8.3 nM; R668A: χ2=0.005, Vmax=0.892, Ki=52 nM; Hill: WT: χ2=0.009, n=0.55, Ki=51 nM; R668A: χ2=0.055, n=0.61, Ki=198 nM. Error bars represent standard error, data points represent the average of three replicates. (D) SEC-SAXS profile of WT (black) and R668A (red) constructs with Rg values calculated by both the Guinier and Porod methods along with Dmax derived from the P(r) function (E). (F) Measurements of the pO subunit D1-D4 distance over the course of six separate microsecond long all-atom MD simulations of WT IDE. (G) Plot of the pO D1-D4 distance vs the D1-D2-D3-D4 COM dihedral angle over the course of the simulation of WT IDE.

Charge-swapping at the IDE-N/C interface mediates conformational dynamics in the closed state

The WT pO subunit remained closed throughout all simulations yet was found to initially relax to a slightly more open state within the first 200 ns, roughly the same time frame it took the WT O subunit to close (Fig. 4F) and, despite not opening, demonstrated a range of D1-D2-D3-D4 dihedral variation similar to that of the O subunits (Fig. 4E). Interestingly, the conformational variation within our simulations increased greatly after this relaxation point was reached (Fig. 4F). This relaxed conformational state consistently displayed an altered IDE-N/C interface compared to the experimentally determined structures that was replicated by the open WT subunits upon closing, leading to a nearly symmetric pO/pO state (Fig. S11). Biochemically, this relaxed IDE-N/C interface makes more sense than the interface observed experimentally. It has previously been demonstrated that the experimental constraints of crystallographic structure determination force IDE to adopt a closed conformation and determination of the open state can only be accomplished once those restraints are removed, i.e. cryoEM (21). These results suggest that both ensemble methods of structure determination may impact the conformation of IDE, however slightly. The IDE-N/C interface is littered with charged residues, yet few interactions are observed in the ensemble structures (Fig. 5A), whereas our MD simulations reveal a complex hydrogen bonding network (Fig. 5B). We observed an extensive network at the D2-D3 interface, with sporadic patches of interactions between D1 and D4 (Fig. S12). Within the D2-D3 network R668 again stands out as a notably key residue. In addition to forming hydrogen bonds with D309 and E381, we observed R668 form π-π interactions with R311 (Fig. 5C). Arginine-mediated π-interactions have been well characterized in other systems for their ability to stabilize interaction interfaces and facilitate conformational changes (34, 35).

Structural basis of closed state conformational dynamics.

(A) IDE-N/C interface previously solved crystal structures (PDB:2G47 shown) shows side chains are ill-positioned for interaction. (B) IDE-N/C interface formed upon open subunit closing in our MD simulations reveals a complex hydrogen bonding network. (C) Heat map showing conformational geometries that were preferentially sampled in our MD simulations by the open subunits upon closing. Insets highlight how the IDE-N/C interface changes to permit interdomain motion. (D) Plot of the O subunit D1-D2-D3-D4 dihedral angle during a subset of a single WT IDE MD simulation after the open-close transition has been completed. Charge-swapping between residues at the IDE-N/C interface is associated with changes in the D1-D2-D3-D4 dihedral. (E) For most of the simulation, D309 interacts with K483 (black), however, this interaction is broken for ∼100 ns, during which D309 instead interacts with R311 (blue) and R668 (orange). (F) For most of the simulation, D426 interacts with K571 (black), yet this interaction is periodically broken, and D426 instead interacts with K425 (green) and K899 (magenta). When these events of charge-swapping coincide with D309 charge-swapping (E), they are associated with a large change in the D1-D2-D3-D4 dihedral angle (D). When they occur alone, the effect on D1-D2-D3-D4 dihedral is smaller. (G) Model for the catalytic cycle of IDE. IDE-N colored orange, IDE-C colored cyan. Insulin colored by chain (A: magenta, B: yellow). Single subunit of the dimer shown for simplicity, second subunit colored gray.

The IDE-N/C interactions we observe in our simulations are not restricted to discrete cognate pairs; rather, we observed that residues within the network periodically swap among several interaction partners. This “charge-swapping” phenomenon allows IDE-N to rotate or slide against IDE-C to adopt multiple conformations and maintain favorable interacting contacts (Fig. 5D). While multiple, concurrent, instances of charge-swapping were associated with rapid changes in the D1-D2-D3-D4 dihedral angle, singular events of charge-swapping did not appear to be significant enough to stimulate large-scale conformational change (Fig. 5E). Instead, substantial conformational changes appear to be caused by multiple charge-swapping events occurring in conjunction. This may explain why we observed no noticeable change in the conformational dynamics of the pO state subunits when comparing the WT and R668A constructs. Consurf analysis reveals that most of the residues forming these interaction networks are highly conserved among IDE homologs (Fig. S13) (36). This indicates that the revealed rotational motion is likely evolutionarily conserved and offers a potential mechanism by which IDE unfolds and repositions bound peptide substrates to degrade amyloid peptides.

Discussion

At its core, cryoEM is an ensemble method of structure determination, averaging tens of thousands of particles together to generate consensus structures. While methodologies have been well-established to separate out discrete conformational classes, proteins that exhibit constant gradients of structural heterogeneity remain problematic, although significant attention has been devoted to the issue in recent years, with varying levels of success (28, 29, 37). Currently, the best approaches use various dimensionality reduction techniques to analyze the principal components of particle heterogeneity. While a substantial innovation, it remains to be seen how the principal components of structural heterogeneity, derived from a series of static snapshots of a particle at the time of freezing, correlate with the real molecular motions of a protein in an aqueous environment. For many systems, especially enzymes of clinical significance, proper characterization of the conformational dynamics and transitional states of the protein are necessary for a full understanding of protein function and lays the foundation for future therapeutic development. MD simulations can provide this information, yet such studies are often met with skepticism by experimental researchers. By integrating MD simulations with experimental techniques, the weaknesses of one technique can be compensated for by the strengths of other techniques to provide comprehensive information about the system of interest. While this integrative approach has proven beneficial across a wide range of systems, including proteorhodopsin, the ribosome, and rotary ATPases, perhaps the best example of this approach is the CRISPR-Cas9 system, where an exhaustive number of studies have combined cryoEM, crystallography, and traditional biophysical techniques with traditional and accelerated MD methods to drive the field forward at an astounding pace (38–43). Here, we contribute to the growing body of literature fusing computational and experimental techniques. We used multibody analysis to reveal the dominant components of structural heterogeneity within our cryoEM particle population, which suggest the conformational dynamics of IDE include a significant rotational component that was not evident from analysis of the ensemble structures. All-atom MD simulations supported the conformational dynamics predicted from cryoEM and offered insight into how promiscuous interactions with a dynamic hydrogen bonding network at the IDE-N/C interface permit the domains to move against one another and suggested that R668 played a key role in mediating IDE-N/C interactions. MD simulations of an R668A mutant revealed significantly altered open-closed dynamics relative to WT. These results were further supported by experimental enzymatic kinetic assays and SEC-SAXS experiments.

IDE utilizes both unfoldase and protease activities to degrade clinically-relevant peptides, including three glucose-regulating hormones with contrary effects – insulin, amylin, and glucagon – and Aβ, the accumulation of which is associated with the progression of Alzheimer’s disease. This has made IDE an attractive target for therapeutic intervention (9, 18). One major challenge is how to better control selectivity when the substrates IDE degrades are highly diverse in sequence and structure. It remains unknown how IDE unfolds substrates prior to cleavage. Our studies characterize the conformational dynamics of IDE and offer insight into the catalytic mechanisms, revealing that IDE-N rotates against IDE-C. We envision that this motion works in concert with the open-close transition to reposition and unfold substrates prior to cleavage. This rotation appears to be largely unconstrained in the open state and may play a role in promoting substrate capture. In the closed state, we found the rotational motion to be mediated by charge-swapping events at the IDE-N/C interface. These events are supplemented by interactions between hydrophobic patches that provide a non-specific interaction interface and allow the domains to slide against one another. We found that IDE-N/C interactions were mediated by the key residue R668, and mutation of this residue altered the enzymatic activity and conformation of IDE, likely by impairing the open-close transition. Our data suggests that, while the R668A mutation globally impaired IDE activity, R668A may be more significantly impaired in its degradation of larger, folded substrates compared to shorter, disordered substrates due to its additional effect on Ki value for insulin. We envision that IDE conformational dynamics can regulate solvent access to the catalytic chamber and substrate binding, unfolding, and cleavage are all intimately linked.

Our results allow us to put forth a refined model describing how IDE recognizes amyloid peptides that are diverse in size and shape (Fig. 5F). The catalytic cycle of IDE starts with at least one subunit of dimeric IDE adopting the open state, exposing the interior of the catalytic chamber to potential substrates (21). The unconstrained motion of IDE-N likely facilitates substrate interaction in a variety of initial orientations. IDE-N and IDE-C have negatively and positively charged surfaces, respectively, and are able to attract peptide substrates with complementary charge profiles (21, 22). Such peptides have a high dipole moment and are often prone to aggregation. However, in the open state, the door subdomain, which contains the key catalytic residues, is highly flexible, rendering the open state catalytically incompetent (21, 24). Proper positioning and unfolding of substrate is necessary to stabilize the cleft prior to catalysis. Substrate binding has been suggested to enhance IDE closing through charge complementarity (44). It is likely that the rotation of IDE-N permits a range of closing pathways to accommodate non-optimal substrate binding geometry (Fig. 3). If the closing geometry permits R668 to interact with E381 or D309, the closing reaction continues to completion. If the closing geometry does not permit the R668-mediated interactions, this could be a signal of improper substrate binding, allowing IDE-N to open and either release improper substrate or reposition for another closing attempt. Once closed, IDE-N is capable of rotating against IDE-C mediated by charge-swapping event at the IDE-N/C interface (Fig. 5). The combined motions associated with the open to closed transition and rotation between IDE-N and IDE-C can lead to the distortion of the secondary structure of bound substrate and facilitate unfolding and proper positioning of substrate for cleavage. Our data indicates that IDE is able to accommodate any possible orientation of insulin tethering within the catalytic site (Fig. 1), suggesting that IDE is able to cut both chains of insulin in rapid succession without requiring significant repositioning (25). IDE then opens to release cleavage products and the cycle begins anew.

Our integrative structural analyses provide the structural basis for IDE conformational dynamics and their roles in IDE catalytic cycle and the framework for future studies. We speculate that multi-cycles of open-close transition and rotation driven by IDE conformational dynamics will be required for its unfoldase activity. However, the kinetic timescale of this reaction is fast, estimated to be ∼500 ms, rendering investigation challenging (25). Furthermore, the mechanism by which IDE opens remains unknown. Future integrative structural analysis for the better understanding and control of IDE catalytic cycle will provide a means to modulate IDE activity in the substrate specific manner to realize its therapeutic potential.

Materials and Methods

Expression and purification of human IDE

Cysteine-free IDE (referred to as IDE in this study) was expressed in E. coli BL21 (DE3) cells (at 25 °C for 20 h, 0.5mM IPTG induction using T7 medium). The R668A mutation was introduced into this construct by site-directed mutagenesis with the following primer pair: CCGGAAATTGTTAAGAGATGCCATATATGCTTCTTTG (forward) CAAAGAAGCATATATGGCATCTCTTAACAACCGGGC (reverse) and verified by sequencing. Recombinant IDE proteins were purified by Ni-NTA, source-Q, and Superdex 200 columns as previously described (21). Protein was aliquoted and stored at -80° C.

CryoEM data collection and analysis

Thawed IDE aliquots were further purified by Superdex 200 chromatography using buffer containing 20 mM HEPES, pH 7.2, 150 mM NaCl, 10mM EDTA and then mixed with FabH11-E at an equal molar ratio. FabH11-E-IDE complex was purified by Superdex 200 chromatography in the presence of insulin at molar ratio IDE:insulin=2:1. Insulin was purchased from SIGMA (91077C). All grids, 300 mesh carbon or gold holey nanowire grids were plasma cleaned with O2 and H2 for 10 secs using a Solarus plasma cleaner (Gatan) and plunged at 133 milli-second using Chameleon, Spotiton-based technology. All images were acquired using a Titan Krios microscope (FEI) operated at 300KeV with a Gatan K3 direct electron detector (Gatan) in counting mode. Images were automatically acquired using Leginon (45). Images were processed using software integrated into RELION (46). Frames were aligned using MotionCor2 (47) software with dose weighting, CTF was estimated using Gctf (48), particles were picked and extracted automatically using RELION. Particle stacks were processed through several rounds of 2D and 3D classification. Selected classes were then processed for high-resolution 3D refinement. The flowchart and detailed data processing is summarized in Figure S1. Finally, the overall map was improved by particle polishing in RELION and sharpening. The final resolution was estimated using Fourier Shell Correlation (FSC=0.143, gold-standard). The density fitting and structure refinement was done using UCSF CHIMERA (49), COOT (50), REFMAC5 (51) and PHENIX (52).

Multibody analysis

We defined three rigid bodies as shown in figure 2A for the analysis. The size of the user-defined bodies has been theorized to play a significant role in the success of multibody refinement, most likely due to the fact that larger bodies will have a stronger signal-to-noise ratio (28). With a Fab bound to each of IDE-N within IDE dimer, we assessed the impact of body size on multibody refinement by examining how density corresponding to 2 Fv regions, 1 Fv region, or no Fv regions would affect the multibody refinement. Fab has a variable region (Fv), and a constant region (Fc) and we chose to subtract and mask out the density corresponding to the Fab Fc, as we speculated that any motion between the Fab Fv and Fc domains would bias the results of the multibody refinement. The quality of the multibody results was assessed based on the subsequent improvement in the Coulomb potential map quality and calculated resolution. In the absence of any Fv density, the maps resulting from multibody refinement had the best calculated resolution, yet the map quality was quite poor overall; much of the density appears globular and featureless, particularly in the IDE-N regions (Fig. S14). Conversely, when Fv density was present on both IDE-N bodies, the resolution of the multibody output maps (4.5 Å) was worse than the resolution of the input map (4.3 Å). Thus, we found that the greatest improvement occurred when the Fv density was present on only the exterior body of the pO subunit. In this case, multibody refinement improved both the calculated resolution and density quality.

Following multibody refinement, the structural heterogeneity for each body was analyzed along six principal component vectors, resulting in 18 vectors describing the primary components of structural variation within the IDE dimer. We analyzed the variation along these vectors as a proxy for protein motion. The results of the multibody analysis indicate the degree of structural variation that is explained by motion along each of the primary component vectors. For all IDE states, we observed no single component vector could explain the majority of structural variance (Table S1-6). Given the underlying assumptions of multibody refinement, there is a question of whether any observed motions are biologically relevant or simply the result of improper particle alignment. We reasoned that the greater the degree of variation explained by motion along a specified component vector, the greater the likelihood was that the motion would be biologically relevant. Therefore, we focus on the top 9 component vectors, which cover ∼75% of total variance for each state, in greater detail.

Molecular dynamics simulations

All-atom MD simulations were prepared from the O/pO structure of IDE (PDB: 7RZK) and necessary CHARMM force field PARAM36 files using QwikMD (53). Gaps in the structure were modeled in using the AlphaFold model of IDE. A three-step minimization-annealing-equilibration process was used to generate an equilibrated system. Simulations were performed under NPT (constant numbers of particles N, pressure P, and temperature T) conditions at 310 K and 1 atm with periodic boundary conditions in NAMD3.0 (54). Explicit solvent was described with the TIP3P model (55). Custom TCL scripts in VMD were used to calculate the D1-D4 centers-of-mass distance and D1-D2-D3-D4 centers-of-mass dihedral angle. Residues 964-988 in D4 that are mostly absence in all static structures of IDE were built using Alpha-fold, which is highly flexible in our MD simulations. They were omitted from the analysis of MD simulations, as the inclusion of this flexible loop significantly altered the center-of-mass independent of global conformational change.

Enzymatic assays

IDE constructs were exchanged into activity buffer comprised of 25 mM Tris-HCl pH 7.5, 150 mM NaCl, and 10 mM ZnCl2. MCA-RPPGFSAFK(Dnp) was purchased from Enzo Life Sciences (BMLP2270001), resuspended in dimethyl sulfoxide at a concentration of 5 mM and diluted to 5 μM with activity buffer. For activity measurements, 5 μM MCA-RPPGFSAFK(Dnp) was mixed with the desired IDE construct at concentrations ranging from 1-100 nM in the presence or absence of 1 mM ATP in a 200 μl reaction. Fluorescence was monitored every 30 sec for 30 minutes with excitation/emission wavelengths of 320/405 nm at 37° C. Initial velocity was calculated during the linear range. Insulin was purchased from MP Biomedicals (#0219390010), resuspended in 0.01N HCl at 1 mM and diluted to desired concentration with activity buffer. For competition assays, 5 nM IDE construct and 5 μM MCA-RPPGFSAFK(Dnp) was mixed with insulin ranging from 0-100 μM. Fluorescence was monitored every 30 sec for 30 minutes with excitation/emission wavelengths of 320/405 nm at 37° C. Initial velocity was calculated during the linear range for each concentration of insulin and normalized relative to the respective construct’s velocity in the absence of insulin to maintain consistency across plates, yielding values for relative inhibition which were plotted vs [insulin] and fit to the Michaelis-Menten equation in xmgrace to generate apparent Ki values. All experiments were performed in triplicate.

Size-exclusion chromatography coupled small-angle X-ray scattering

SAXS/WAXS data collection was employed the Life Sciences X-ray Scattering (LIX) Beamline at the National Synchrotron Light Source II (NSLS II) at Brookhaven National Laboratory in Upton, NY. Briefly, 60ul of samples in solution were pipetted into PCR tubes, placed into a Bio-Inert Agilent 1260 Infinity II HPLC multisampler and measured using the isocratic SEC-SAXS format at the beamline (56). 50ul of sample was injected into the Phenomenex Biozen dSEC2 3 μm bead size with 200Å pore size column was utilized at a flow rate of 0.35mL/min for 25 minutes. SAXS and WAXS data are collected simultaneously on a Pilatus 1M (SAXS) and Pilatus 900K (WAXS) detectors with a 2 second exposure (57). Data from both detectors is then scaled and merged. Intensity is normalized using the water peak height at 2.0 Å -1. Data processing and analysis was performed using py4xs and lixtools in jupyter lab. Buffer frames 100-125 were averaged and used for subtraction of averaged frames under the peak of interest.

eLife Assessment

The manuscript by Mancl et al. provides valuable mechanistic insights into the conformational dynamics of Insulin Degrading Enzyme (IDE), a zinc metalloprotease involved in the clearance of various bioactive peptides. Supported by a convincing combination of cryo-EM, SEC-SAXS, enzymatic assays, and molecular dynamics simulations, the study characterizes the dynamic transitions between IDE's open and closed states in the presence of a sub-saturating concentration of insulin. This work contributes to a refined model of IDE's functional cycle, enhancing our understanding of its role in proteolysis.

Reviewer #1 (Public review):

Summary:

Mancl et al. present cryo-EM structures of the Insulin Degrading Enzyme (IDE) dimer and characterize its conformational dynamics by integrating structures with SEC-SAXS, enzymatic activity assays, and all-atom molecular dynamics (MD) simulations. They present five cryo-EM structures of the IDE dimer at 3.0-4.1 Å resolution, obtained with one of its substrates, insulin, added to IDE in a 1:2 ratio. The study identified R668 as a key residue mediating the open-close transition of IDE, a finding supported by simulations and experimental data. The work offers a refined model for how IDE recognizes and degrades amyloid peptides, incorporating the roles of IDE-N rotation and charge-swapping events at the IDE-N/C interface.

Strengths:

The study by Mancl et al. uses a combination of experimental (cryoEM, SEC-SAXS, enzymatic assays) and computational (MD simulations, multibody analysis, 3DVA) techniques to provide a comprehensive characterization of IDE dynamics. The identification of R668 as a key residue mediating the open-to-close transition of IDE is a novel finding, supported by both simulations and experimental data presented in the manuscript. The work offers a refined model for how IDE recognizes and degrades amyloid peptides, incorporating the roles of IDE-N rotation and charge-swapping events at the IDE-N/C interface. The study identifies the structural basis and key residues for IDE dynamics that were not revealed by static structures.

Weaknesses:

Based on MD simulations and enzymatic assays of IDE, the authors claim that the R668A mutation in IDE affects the conformational dynamics governing the open-closed transition, which leads to altered substrate binding and catalysis. The functional importance of R668 would be substantiated by enzymatic assays that included some of the other known substrates of IDE than insulin such as amylin and glucagon.

It is unclear to what extent the force field (FF) employed in the MD simulations favors secondary structures and if the lack of any observed structural changes within the IDE domains in the simulations - which is taken to suggest that the domains behave as rigid bodies - stems from bias by the FF.

Reviewer #2 (Public review):

Summary:

The manuscript describes various conformational states and structural dynamics of the Insulin degrading enzyme (IDE), a zinc metalloprotease by nature. Both open and closed-state structures of IDE have been previously solved using crystallography and cryo-EM which reveal a dimeric organization of IDE where each monomer is organized into N and C domains. C-domains form the interacting interface in the dimeric protein while the two N-domains are positioned on the outer sides of the core formed by C-domains. It remains elusive how the open state is converted into the closed state but it is generally accepted that it involves large-scale movement of N-domains relative to the C-domains. The authors here have used various complementary experimental techniques such as cryo-EM, SAXS, size-exclusion chromatography, and enzymatic assays to characterize the structure and dynamics of IDE protein in the presence of substrate protein insulin whose density is captured in all the structures solved. The experimental structural data from cryo-EM suffered from a high degree of intrinsic motion among the different domains and consequently, the resultant structures were moderately resolved at 3-4.1 Å resolution. A total of five structures were generated by cryo-EM. The authors have extensively used Molecular dynamics simulation to fish out important inter-subunit contacts which involve R668, E381, D309, etc residues. In summary, authors have explored the conformational dynamics of IDE protein using experimental approaches which are complemented and analyzed in atomic details by using MD simulation studies. The studies are meticulously conducted and lay the ground for future exploration of the protease structure-function relationship.

Author response:

The following is the authors’ response to the original reviews.

Reviewer #1 (Public review):

Summary:

Mancl et al. present cryo-EM structures of the Insulin Degrading Enzyme (IDE) dimer and characterize its conformational dynamics by integrating structures with SEC-SAXS, enzymatic activity assays, and all-atom molecular dynamics (MD) simulations. They present five cryo-EM structures of the IDE dimer at 3.0-4.1 Å resolution, obtained with one of its substrates, insulin, added to IDE in a 1:2 ratio. The study identified R668 as a key residue mediating the open-close transition of IDE, a finding supported by simulations and experimental data. The work offers a refined model for how IDE recognizes and degrades amyloid peptides, incorporating the roles of IDE-N rotation and charge-swapping events at the IDE-N/C interface.

Strengths:

The study by Mancl et al. uses a combination of experimental (cryoEM, SEC-SAXS, enzymatic assays) and computational (MD simulations, multibody analysis, 3DVA) techniques to provide a comprehensive characterization of IDE dynamics. The identification of R668 as a key residue mediating the open-to-close transition of IDE is a novel finding, supported by both simulations and experimental data presented in the manuscript. The work offers a refined model for how IDE recognizes and degrades amyloid peptides, incorporating the roles of IDE-N rotation and chargeswapping events at the IDE-N/C interface. The study identifies the structural basis and key residues for IDE dynamics that were not revealed by static structures.

Weaknesses:

Based on MD simulations and enzymatic assays of IDE, the authors claim that the R668A mutation in IDE affects the conformational dynamics governing the open-closed transition, which leads to altered substrate binding and catalysis. The functional importance of R668 would be substantiated by enzymatic assays that included some of the other known substrates of IDE than insulin such as amylin and glucagon.

We have included amyloid beta in our enzymatic assays, as shown in Figure 5D, and have updated the manuscript text accordingly. The R668A mutation results in a loss of dose-dependent competition with amyloid beta, but not with insulin. To further substantiate this unexpected finding, we plan to undertake a comprehensive biochemical characterization of the R668A mutation across a variety of substrates, followed by structural analysis of this mutant. However, these investigations are beyond the scope of the current study and, if successful, warrant a separate publication.

It is unclear to what extent the force field (FF) employed in the MD simulations favors secondary structures and if the lack of any observed structural changes within the IDE domains in the simulations - which is taken to suggest that the domains behave as rigid bodies - stems from bias by the FF.

We utilized the widely adopted CHARMM36 force field, whose parameters have been validated by thousands of previous studies. As shown in Figure 2A, our simulations reveal small but noticeable fluctuations in intradomain RMSD values. However, after careful examination, we found that these changes do not correspond to any biologically meaningful motions based on previously reported structural and biophysical characterizations of IDE (e.g., Shen et al., Nature 2006; Noinaj et al., PLOS One 2011; McCord et al., PNAS 2013; Zhang et al., eLife 2018, and references therein).

Reviewer #2 (Public review):

Summary:

The manuscript describes various conformational states and structural dynamics of the Insulin degrading enzyme (IDE), a zinc metalloprotease by nature. Both open and closed-state structures of IDE have been previously solved using crystallography and cryo-EM which reveal a dimeric organization of IDE where each monomer is organized into N and C domains. C-domains form the interacting interface in the dimeric protein while the two N-domains are positioned on the outer sides of the core formed by Cdomains. It remains elusive how the open state is converted into the closed state but it is generally accepted that it involves large-scale movement of N-domains relative to the C-domains. The authors here have used various complementary experimental techniques such as cryo-EM, SAXS, size-exclusion chromatography, and enzymatic assays to characterize the structure and dynamics of IDE protein in the presence of substrate protein insulin whose density is captured in all the structures solved. The experimental structural data from cryo-EM suffered from a high degree of intrinsic motion among the different domains and consequently, the resultant structures were moderately resolved at 3-4.1 Å resolution. A total of five structures were generated by cryo-EM. The authors have extensively used Molecular dynamics simulation to fish out important inter-subunit contacts which involve R668, E381, D309, etc residues. In summary, authors have explored the conformational dynamics of IDE protein using experimental approaches which are complemented and analyzed in atomic details by using MD simulation studies. The studies are meticulously conducted and lay the ground for future exploration of the protease structure-function relationship.

Reviewer #1 (Recommendations for the authors):

The manuscript reads well, however, there are minor details throughout that would tighten it up and, in some cases, make it easier to approach for a broader readership:

Abstract

(1) R668 is referred to by its one-letter code throughout the main text but referred to as arginine-668 in the abstract. The abstract should be corrected to R668.

This has been corrected.

(2) The authors should consider reordering the significance of their work as it is listed at the end of the abstract. As the work first and foremost "offers the molecular basis of unfoldase activity of IDE and provides a new path forward towards the development of substrate-specific modulators of IDE activity" these should come before "the power of integrating experimental and computational methodologies to understand protein dynamics".

We have revised abstract substantially to incorporate the new findings. Consequently, the sentence for "the power of integrating experimental and computational methodologies to understand protein dynamics" has been removed.

Main text

(1) Cryo-EM is consistently referred to as cryoEM throughout the text. The commonly accepted format for referring to cryogenic electron microscopy is cryo-EM. The authors are asked to consider revising the text accordingly.

The text has been revised.

(2) Introduction: The authors are asked to consider including a figure (panel) that provides the general reader with an overview of IDE architecture and topology as a point of reference in the introduction to understanding the pseudo symmetry in IDE, domains, and IDE-C relative to IDE-N, etc. This is relevant for reading most of the figures.

We have added a new figure 1 to provide the background and questions to be answered.

(3) The authors should consider renaming some of the headers in the results section to include the main conclusion. For instance, "CryoEM structures of IDE in the presence of a sub-saturating concentration of insulin" is not really helpful for the reader to understand the work, while "R668A mediates IDE conformational dynamics in vitro" is.

The headings have been altered in an effort to be more informative.

(4) It is unclear what the timescale for insulin cleavage is for IDE. Clearly, it is possible for the authors to capture an insulin-bound IDE from within the 7 million particles, but what is the chance of this? The authors emphasize the IDE:insulin ratio relative to previous experiments, but surely the kinetics would be the same in the two experiments that were presumably set up exactly the same way. In the context of this, the authors should disclose how concentrations were estimated experimentally. The authors are encouraged to touch upon the subject of time scales to tie up cryo-EM and enzyme experiments with MD simulations.

Both reviewers posted the question about time-scale relevant to IDE catalysis. In response to this request, we have revised the manuscript to address the relevance of key kinetic timescales. Specifically, we now discuss the open/closed transition (~0.1 second) and insulin cleavage (~2/sec), both established experimentally in prior studies (McCord et al PNAS 2013).

IDE concentrations were determined by spectrometry (Nanodrop and/or Bradford assay), and its purity was confirmed to be greater than 90% by SDS-PAGE. Insulin was purchased commercially, weighed, and dissolved in buffer, with its concentration subsequently verified using Nanodrop. Catalytically inactive IDE and insulin were mixed and incubated for at least 30 minutes. Given IDE’s low nanomolar affinity for insulin, and the sub-stoichiometric insulin concentrations used, sufficient time was allowed for insulin to bind IDE and remain bound.

To distinguish between IDE’s unfoldase and protease activities, all structural analyses were performed in the presence of EDTA, which chelates catalytic zinc, thereby inactivating IDE. This approach inhibits the enzyme’s catalytic cycle and allows us to capture the fully unfolded state of insulin bound to IDE in its closed conformation, representing the endpoint of the reaction. Under these conditions, the only meaningful kinetic parameter available for investigation was the unfolding of insulin by IDE.

To elaborate the interaction between IDE and insulin in the catalytically relevant time regime, we investigated IDE–insulin interactions within the millisecond time regime by rapidly mixing IDE with a large molar excess of insulin for approximately 120 milliseconds for the cryo-EM single particle analysis. Under these conditions, we observed that both IDE subunits in the dimer predominantly adopt open states, which are distinct from those previously reported. This observation suggests a potential mechanism of allostery in IDE function.

(5) It should be included in the main text that the data was processed with C1 symmetry and not just in Table 1. This is more useful information for understanding the study than the number of micrographs.

We have stated that the data was processed with C1 symmetry at the start of the results section.

(6) The authors should consider adding speculation on what the approximately 6 million particles that did not yield a high-resolution structure represent.

In cryo-EM single particle analysis, particle selection is typically performed automatically using software such as Relion. Due to the low signal-to-noise ratio, many “junk particles”—originating from contaminants such as ice, impurities, aggregates, or incomplete particles—are inevitably included along with the particles of interest. It is standard practice to filter out these junk particles during data processing. In our case, we estimate that the majority of the 6 million particles are likely junk. However, we cannot fully exclude the possibility that some of these particles may originate from IDE and carry potentially useful information about its conformational heterogeneity. Nonetheless, current cryo-EM single particle analysis methods face significant challenges in objectively recovering and interpreting such particles.

Reviewer #2 (Recommendations for the authors):

I have some minor comments regarding the manuscript which are given below.

(1) For O/O state, it will be great to see an explanation regarding why the values are dissimilar for 0.5 and 0.143 FSC.

All of our IDE structures (including previously published data) demonstrate a dip/plateau at moderate resolution in their FSCs. We interpret this an indicator of structural heterogeneity, as the dip/plateau is smallest in the pC/pC state, becomes larger when one of the subunits is open, and is largest when both subunits are open. Because both subunits within the O/O state are highly heterogeneous, the FSC dipped below the 0.5 threshold. Other states, such as the O/pO, display the same FSC trend, the dip remains slightly above the 0.5 threshold.

(2) O/pO state is moderately resolved at 4.1 Å, but this state is populated with many particles (328,870). Can the resolution be improved by more extensive sorting of heterogenous particles which intrinsically causes misalignment amongst particles?

Unfortunately, no. As shown by the local resolution maps in Figure 1-figure supplement 1, the primary source of misalignment is the IDE-N region in the open subunit. We have found that IDE-N is nearly unconstrained in its conformational flexibility in the open state, and does not appear to adopt discrete states, our attempts to better classify particles have failed. We speculate that this may be a failing in kmeans cluster based classification, and this is part of the driving force behind our exploration of advanced methods of heterogeneity analysis.

(3) Given the observation that capturing a substrate-bound open state is difficult, it can be assumed that the substrate capture in the catalytic cleft is a fast event. Please comment on the possible time frame of unfolding of substrate and catalysis. Can authors comment on any cryo-EM experiments that can deal with such a short time frame? If there is a possibility to include data from such experiments, then it may be considered.

This has been addressed in conjunction with the previous reviewer’s comment (see above). Specifically, we now discuss the open/closed transition (~0.1 second) and insulin cleavage (~2/sec), both established experimentally in prior studies. Additionally, we investigated IDE–insulin interactions by rapidly mixing IDE with a large molar excess of insulin for approximately 120 milliseconds for the cryo-EM single particle analysis. Under these conditions, we observed that both IDE subunits in the dimer predominantly adopt open states, which are distinct from those previously reported. This observation suggests a potential mechanism of allostery in IDE function.

(4) How long was incubation time after adding any substrates, such as insulin? Can different incubation times be tested to generate additional information regarding other conformational states that lie in between open and closed states?

The incubation time for IDE with insulin prior to cryo-EM grid freezing was approximately 30 minutes. We agree that it would be exciting to explore shorter time frames to identify new conformational states. As discussed above, we have rapidly mixed IDE with a large molar excess of insulin for approximately 120 milliseconds for the cryo-EM single particle analysis. Under these conditions, we observed that both IDE subunits in the dimer predominantly adopt open states, which are distinct from those previously reported. This observation suggests a potential mechanism of allostery in IDE function.

(5) A complex network of hydrogen bonding interaction initiated by R668 latching onto N-domain is mentioned in MD simulation studies but it is not clear why cryo-EM experiments did not capture such stabilized structures.

We believe that two main factors have prevented us from observing the hydrogen bonding network in our cryo-EM structures. The first factor is the requirement to freeze the sample in liquid ethane. According to the second law of thermodynamics, lowering the temperature reduces the effect of entropy. Our findings suggest that residue R668 interacts with several neighboring residues through a network of polar and electrostatic interactions, rather than being limited to a single partner. These interactions facilitate both the open-closed transitions and rotational movements between IDE-N and IDE-C. From a thermodynamic perspective, these interactions have both enthalpic and entropic components, and cooling the sample diminishes the entropic contribution. In line with this, we observe that the closed-state domains in our cryo-EM studies are positioned closer together than in our MD simulations, though not as tightly as in crystal structures of IDE. This implies that cryogenic data collection may constrain the interface between IDE-N and IDE-C, which can further alter the equilibrium for the network of R668 mediated interactions.

Secondly, our cryo-EM structures represent ensemble averages of tens to hundreds of thousands of particles. MD simulations indicate that IDE-N and IDE-C can rotate relative to one another, resulting in considerable variability in residue interactions. However, the level of particle density in our cryo-EM data does not permit sufficiently fine classification to resolve these differences. As a result, distinct hydrogen bonding networks are likely averaged out in the ensemble structure, particularly in the case of R668, which is indicated to interact with multiple neighboring residues in the conformation-dependent manner. This averaging effect may also contribute to our inability to achieve resolutions below 3 Å.

(6) Despite the observation that IDE is an intrinsically flexible protein, it seems probable that differently-sized substrates might reveal additional interaction networks formed by other novel key players apart from just R668. Will it be helpful to first try this computationally using MD simulations and then try to replicate this in cryo-EM experiments? If needed, additional simulation time may be added to the MD analysis. Please comment!

We agree that this is an exciting avenue to explore. Doubly so when considered in light of our R668A enzymatic results with amyloid beta. However, several challenges must be overcome before we can explore this direction effectively:

(1) We lack experimental knowledge of the initial interaction event between IDE and substrate. All substrate-bound IDE structures have been obtained after unfolding and positioning for cleavage has occurred. Without a solid foundational model for the initial interaction event between IDE and substrate, the interpretation of subsequent MD simulations is open to question.

(2) We have previously observed minimal effect of substrate on IDE in all-atom MD simulations. We believe that observable effects would require a much longer time scale than is currently achievable with all-atom MD, so have turned to Upside, a coarse-grained method to overcome these limitations, but Upside handles side chains with presumptive modeling, which prevent the identification of potential novel residue interactions.

(3) Due to the conformational heterogeneity present within IDE cryo-EM datasets, we struggle to obtain sufficient resolution to clearly identify side chain interactions at the domain interface (see response to 5).

Given these challenges, we plan to explore these directions in future manuscripts.

(7) What is the possibility of water interaction networks and dynamism in this network to contribute to the overall dynamics of the protein in the presence and absence of substrates? How symmetric these networks be in the four domains of dimeric IDE?

This is an interesting idea that we have begun to explore, but consider to be outside the scope of this work. Currently, we do not have any MD simulations containing substrate with explicit solvent (Upside uses implicit solvent), and solvent atoms were removed from our all-atom simulations prior to analysis to speed up processing. That being said, preliminary WAXS data suggests that there may be a difference in water interaction interfaces between WT and R668A IDE, and this is a lead we plan to pursue in future work.

(8) Line 214: Please fix the typo which wrongly describes closed = pO.

This is not a typo, but it is confusing. The pO state has previously been defined as the closed state of IDE lacking bound substrate as determined by cryo-EM. This differentiates the pO state from the pC state, where the pC state contains density indicative of bound substrate. As the MD simulations were conducted with the apo-state, the closed state the simulations were initialized from was the pO state structure, which represents the substrate-free closed state as determined by cryo-EM. We realize that this difference is probably unnecessary to the majority of readers, and have removed the (pO) specificity to avoid confusion.

(9) It is not clear why a cryo-EM structure was not attempted for the R668A mutant. If the authors have tried to generate such a structure, it should be mentioned in the manuscript. Such a structure should yield more information when compared to SAXS experiments.

We have not attempted to obtain a cryo-EM structure for the R668A mutant. Our SAXS analysis suggests a transition from a dominant O/pO state to a dominant O/O state. The O/O state is known to exhibit the highest degree of conformational heterogeneity, which severely limits structural insights. We are working to better handle the sample preparation of IDE and perform such analysis without the need to use Fab. We plan to further characterize IDE R668A biochemically and potentially explore other mutations that would provide insights in how IDE works. Armed with that, we will perform the structural analysis of such IDE mutant(s).

Version 2

Abstract

Insulin degrading enzyme (IDE) is a dimeric M16A zinc metalloprotease that degrades amyloidogenic peptides diverse in shape and sequence, including insulin and amyloid-β, to prevent toxic amyloid fibril formation. IDE has a hollow catalytic chamber formed by two ∼55 kDa N- and C-domains (IDE-N and IDE-C, respectively), in which peptides bind, unfold, and are repositioned for proteolysis. IDE is known to transition between a closed state, poised for catalysis, and an open state, able to release cleavage products and bind new substrate. Here, we present five cryo-EM structures of the IDE dimer at 3.0-4.1 Å resolution, obtained in the presence of a sub-saturating concentration of insulin. Combining cryo-EM heterogeneity analysis with all-atom molecular dynamics (MD) simulations, we identified the structural basis and key residues for IDE conformational dynamics that were not revealed by IDE static structures. Notably R668 serves as a molecular latch mediating the open-close transition and facilitates key protein motions through charge-swapping interactions at the IDE-N/C interface. Our small-angle X-ray scattering and enzymatic assays of an R668A mutant indicate a profound alteration of conformational dynamics and catalytic activity. By integrating coarse-grained Upside MD simulations, our analysis reveals that IDE unfolds its substrates through the coordinated motion between IDE-N and IDE-C, as well as β-sheet formation between IDE and insulin. Additionally, our time-resolved cryo-EM analysis uncovers IDE allostery within the IDE dimer. Collectively, our findings demonstrate the strength of combining experimental and computational approaches to probe protein dynamics and pave the way for developing substrate-specific modulators of IDE activity.

Introduction

Protein homeostasis (a.k.a. proteostasis) is maintained by three primary mechanisms: chaperones, ubiquitination/proteasome, and autophagy (Hipp et al, 2019; Lopez-Otin et al, 2013). Disruptions in proteostasis can lead to the accumulation of amyloid fibrils and subsequent human diseases (Chiti & Dobson, 2006; Chiti & Dobson, 2017; Greenwald & Riek, 2010). As a result, many proteases have evolved to specifically target amyloidogenic peptides (Malito et al, 2008; Saido & Leissring, 2012). Among those, the cryptidase family, which includes the M16 metalloproteases insulin degrading enzyme (IDE) use an internal catalytic chamber, or “crypt”, to capture and selectively degrade the monomeric form of amyloid peptides to control the formation of amyloid fibrils (Liang et al, 2022; Malito et al., 2008). IDE effectively degrades various bioactive peptides, including amyloid-β (Aβ), a peptide associated with the progress of Alzheimer’s disease, and three blood glucose-regulating hormones: insulin, amylin, and glucagon (Tang, 2016). Consequently, defects in IDE alter the progression of type 2 diabetes mellitus and Alzheimer’s disease in animal models and are linked to these diseases in humans (Bertram et al, 2000; Bertram et al, 2007; Farris et al, 2003; Farris et al, 2004; Kim et al, 2007; Sladek et al, 2007; Tang, 2016; Tanzi & Bertram, 2005; Zeggini et al, 2007). IDE is a promising therapeutic target, as its inhibition improves glucose tolerance, yet progress has been hampered by the contrary actions of its diverse substrate pool (Maianti et al, 2014; Maianti et al, 2019; Tang, 2016). For example, IDE overexpression has been shown to reduce Aβ loads, in mice, however, this reduction is accompanied by hyperglycemia from a similar reduction in insulin levels (Leissring et al, 2003).

IDE-mediated substrate recognition and degradation is a complex process. IDE is capable of degrading substrates with diverse sequences and structures, such as Aβ and the three glucose-regulating hormones: insulin, amylin, and glucagon (Malito et al., 2008; Tang, 2016). Despite this diversity, IDE exhibits high selectivity—for example, it preferentially degrades insulin and atrial natriuretic peptide over insulin-like growth factor-1 and brain natriuretic peptide, respectively (Malito et al., 2008). The presence of ATP, inositol phosphate, or specific mutations can markedly enhance the catalytic rate for small peptide substrates like bradykinin, but does not significantly affect the degradation of larger substrates such as insulin and Aβ (Im et al, 2007; McCord et al, 2013; Song et al, 2017; Song et al, 2004). Additionally, the binding location of IDE inhibitors within the catalytic chamber can influence substrate preference, making IDE more likely to degrade one substrate over another (Charton et al, 2014; Maianti et al., 2019). This complexity arises partly from the interplay between IDE’s unfoldase and protease activities, which are essential for degrading diverse substrates (Malito et al., 2008; Tang, 2016; Zhang et al, 2018). IDE relies on its unique large catalytic chamber and associated conformational dynamics to unfold and degrade amyloidogenic substrates. A deeper understanding of the molecular mechanisms by which IDE utilizes its catalytic chamber for substrate unfolding and degradation is crucial for effectively harnessing this enzyme in future therapeutic development.

IDE is a dimeric 110 kDa metalloprotease. Each protomer is comprised of four homologous subdomains organized into two ∼55 kDa domains, IDE-N and IDE-C, connected by a short flexible linker (Li et al, 2006; Shen et al, 2006) (Fig. 1A). Our previous cryo-EM analysis revealed the major conformational states of IDE. In the absence of substrate, each protomer was found to adopt an open (O) or partial open (pO) state which are primarily differentiated by the displacement of IDE-N relative to IDE-C, the combination of which led to three dimeric conformations (O/O, O/pO, pO/pO) (Zhang et al., 2018). It was observed that substrate binding induced both protomers of IDE to fully close, referred to as the partial closed (pC, pC/pC for the dimer) state, as the cryo-EM closed state was found to be slightly more open than previously solved crystal structures, likely influenced by the constraints of the crystal lattice (McCord et al., 2013; Shen et al., 2006). Only the O state permits substrate access to the catalytic chamber, yet the catalytic cleft is stabilized only in the pC state with substrate bound, requiring IDE to undergo a substantial open-close transition during catalysis, the specifics of which are unknown (Zhang et al., 2018).

Overview of structure and function of IDE.

(A) Overall structure of IDE. IDE is comprised of four structurally homologous domains: D1 (red, residues 43-285), D2 (orange, residues 286-515), D3 (blue, residues 545-768), and D4 (cyan, residues 769-1,016). The domains are arranged in two roughly hemispherical regions that enclose the catalytic chamber referred to as IDE-N (comprised of the D1 and D2 domains) and IDE-C (comprising the D3 and D4 domains), which are joined by a linker region (residues 516-544). (B) Primary sequence and overall structure of insulin. Insulin cleavage by IDE as revealed by mass spectrometry (Manolopoulou et al., 2009). Cleavage sites are marked in arrows, which occur within α-helical regions and requires the peptide to be unfolded prior to cleavage. (C) Overview of the IDE catalytic cycle with key questions addressed in this study.

Biochemical and structural analyses indicate that many IDE substrates, such as insulin, must first be unfolded prior to cleavage (Fig. 1B, C). The conformational dynamics of IDE are thought to play a central role in substrate recognition and degradation, yet several mechanistic details underlying IDE’s unfoldase activity remain unresolved (Fig. 1). We hypothesize that the unfoldase and protease activities of IDE are regulated by the relative motions of the IDE-N and IDE-C domains, their interactions with substrates, and allosteric communication between subunits within the IDE dimer. To investigate this, we solved cryo-EM structures of IDE bound to insulin at a 2:1 IDE:insulin ratio, as well as structures captured following rapid mixing of IDE with insulin under functionally relevant conditions. Utilizing newly developed computational methods, we analyzed particle heterogeneity in these structures to infer IDE’s conformational dynamics. By integrating these experimental data with all-atom and coarse-grained molecular dynamics (MD) simulations, we identified and modulated specific molecular interactions to manipulate IDE’s key conformational dynamics, ultimately enhancing our understanding how to control its enzymatic activity in vitro.

Results

Cryo-EM reveals new IDE structures in the presence of a sub-saturating concentration of insulin

We have previously used cryo-EM to solve the structure of IDE in the presence of a 5-fold molar excess of insulin, resulting in a dominant pC/pC state with both substrate binding chambers yielding density corresponding to unfolded insulin (Zhang et al., 2018). To explore the mechanism of IDE-directed substrate unfolding under more physiologically relevant conditions, here we investigated IDE-insulin interactions with an IDE:insulin molar ratio of 2:1. We collected a dataset of ∼7,600 micrographs on a Titan Krios operated at 300 keV, from which 7.2 million particles were picked for processing in RELION. Following an established workflow, and processing the data with C1 symmetry, we generated 5 structures: three previously observed structures with improved resolution (pC/pC, O/O, and O/pO at 3.0 Å, 3.8 Å, and 4.1 Å resolution) and two novel states (O/pC and pO/pC at 3.4 Å and 3.3 Å resolution, respectively) (Fig. 2A, Figure 2 – figure supplement 1, Supplementary file 1). Interestingly, the pC/pC was still found to be dominant, despite IDE and insulin being present at a 2:1 molar ratio, consistent with the allostery of IDE (McCord et al., 2013; Zhang et al., 2018).

Cryo-EM structures.

(A) Overview of the cryo-EM structures. See figure S1 for processing details. (B) Comparison of the open (O), partial open (pO), and partial closed (pC) subunit states present in our cryo-EM structures with domain organization. The distance between the D1 and D4 domain centers-of-mass (D1-D4 COM) along with the dihedral angle formed by the D1-D2-D3-D4 domain centers-of-mass (D1-D2-D3-D4 dihedral) described in Zhang et al. (21) and depicted in Figure 3 – figure supplement 1 were used as biologically important criteria to quantify observed conformations. (C) Insulin density and corresponding model in our cryo-EM structures. Both the A chain (magenta) and B chain (yellow) can fit the density in the exosite and catalytic cleft.

The individual protomers of IDE adopt the same three conformations: O, pO, and pC states, reported previously but there are significant differences in the O and pO states (Fig. 2B) (Zhang et al., 2018). While the pC/pC state is nearly identical to the two states reported previously (RMSD of ∼0.6 Å; PDB ID: 6BFC, 6B3Q) (Zhang et al., 2018), the O and pO states appear more closed when insulin is present, suggesting that the presence of insulin promotes the open to closed transition. Specifically, the pO and O subunits are 3-5 Å more closed based on the distance between the centers-of-mass (COM) of the IDE D1 and D4 subdomains, while the dihedral angle formed by the COM of the D1-D2-D3-D4 subdomains is reduced by 5-10° (Fig. 2B). Alignment of the new O/O and O/pO states to their previously solved counterparts reveals little difference, with a global RMSD of 1.316 Å (O/O, PDB ID: 6B7Y) and 1.405 Å (O/pO, PDB ID: 6BF8), respectively (Zhang et al., 2018). The primary source of variation between the different states remains the degree of opening between the two domains of each subunit, as a global alignment of IDE-N and IDE-C across all structures reveal the domains primarily act as rigid bodies (RMSD <1.4 Å). Consistent with previous observations, the door region, which contains the catalytic zinc binding site, exhibited higher B-factors in the O states than in either the pO or pC state (Zhang et al., 2018) (Figure 2 – figure supplement 2).

All five cryo-EM structures presented here contain four subunits that have clear density present at the catalytic cleft and exosite indicative of bound substrate, thus adopting the pC state. Previous structures have modeled the same chain of insulin in both the exosite and catalytic cleft, but the best models we could build into our density suggest that the chain bound to the exosite is not the same chain positioned for cleavage at the catalytic cleft (Fig. 2C). The preponderance of data suggests it to be unlikely that insulin adopts a preferred orientation within the catalytic chamber, rather the catalytic chamber can accommodate insulin in four different orientations, with either the A or B chain bound to the exosite and catalytic cleft in cis and trans orientations. This promiscuity of binding has been theorized as a reason for why there is no chain preference for the initial insulin cleavage event (Manolopoulou et al, 2009). Crystal structures of insulin are highly compact and globular, with little spatial separation between the N- and C-termini of each chain, and the IDE cleavage sites are located within α-helices (Baker et al, 1988). In its crystallized state, insulin cannot interact with both the exosite and catalytic cleft unless it unfolds, at least partially. Recent work used MD simulations to study the conformational dynamics of insulin in solution and identified several major “elements of disorder” to describe the partially unfolded intermediate structures they observed (Busto-Moner et al, 2021). We docked these structures of insulin into the closed catalytic chamber of IDE and found that the chamber was able to easily accommodate the positioning of either chain near the exosite and catalytic cleft in cis and trans orientations (Figure 2 – figure supplement 3).

Heterogeneity analysis reveals 2 dominant components of structural variability with the IDE cryo- EM particle population

Numerous structures of IDE exist with subunits adopting either an open or closed conformation, yet we lack an understanding of how IDE transitions from an open to a closed state. The simplest explanation of this transition, based on structural data, would be a direct, rigid body translation of IDE-N relative to IDE-C, and this is the predominant model within the field (Shen et al., 2006; Zhang et al., 2018). To better understand what this transition would look like, we measured and plotted the changes in the D1-D4 COM distance versus the changes in the D1-D2-D3-D4 COM dihedral angle for all available cryo-EM structures of IDE (Figure 3 – figure supplement 1). We found that in the absence of substrate, the transition pathway produced a linear relationship between states. However, when our structures generated in the presence of insulin were analyzed, the linear relationship no longer held (Figure 3 – figure supplement 1), suggesting that the simple linear transition typically presumed for the open-close transition between distinct IDE states does not accurately depict the complexity of IDE dynamics.

Conformational dynamics of IDE implied by structural heterogeneity.

(A) All-atom MD simulations analysis. The primary source of structural variance (RMSD) results from the IDE-N moving against IDE-C as a rigid body. Rigid bodies were defined as colored for multibody refinement in RELION. (B-C) Multibody analysis. The range of conformational variance described by the top principal component vectors displays an unexpectedly high degree of rotational motion, as measured by the change in D1-D2-D3-D4 dihedral angle across each vector’s gradient of structural heterogeneity, in both the absence (B) and presence (C) of insulin compared to the expected open-close transition pathway predicted from a linear interpolation of the experimentally determined structures of IDE (dashed line, Figure 3 – figure supplement 1). Two dominant components of structural variance are revealed from multibody analysis: (D) where IDE-N swings relative to IDE-N about the inter-domain linker, and (E) where IDE-N rotates against IDE-C. Starting (orange) and ending (red) states of IDE-N shown with pathway depicted by arrows. IDE-C shown as gray surface.

Recently, several approaches have been developed to understand the conformational heterogeneity present within cryo-EM data. We employed multibody analysis in RELION and 3D variability analysis (3DVA) in cryoSPARC to investigate the conformational heterogeneity within our particle populations (Nakane et al, 2018; Punjani & Fleet, 2021). Of these, the range of motion predicted by multibody analysis most closely matched the reported structures while 3DVA exhibited consistent results across a smaller magnitude of transition (Figure 3 – figure supplements 1-3). Therefore, we focused on multibody analysis below. Multibody analysis models the structural heterogeneity in our data as the result of motions of independent, user-defined rigid bodies. As discussed above, the primary source of structural variation in our structures is the positioning of IDE-N relative to IDE-C. We see little structural changes within the domains, suggesting that IDE domains behave as three rigid bodies, IDE-N in two different subunits within IDE dimer (IDE-N A, IDE-N B) and the dimeric IDE-C (IDE C/C). This is consistent with the structural analysis of M16 family of metalloproteases (Liang et al., 2022). This assertion is further supported by our all-atom molecular dynamics (MD) simulations. While we observed small yet noticeable conformational changes within IDE-N A, IDE-N B, and IDE-C/C, the overall RMSD of entire IDE dimer is significantly higher than those three individual units (Fig. 3A). Multibody analysis defines the principal components of variance along discrete eigenvectors, which we interpret as proxies for the dominant components of molecular motion. We quantified the change in D1-D4 COM distance and D1-D2-D3-D4 COM dihedral of representative structures along the heterogeneity gradients of the top 9 eigenvectors for each of our IDE structures, representing ∼70-90% of the total structural heterogeneity per structure. The results indicated that the particles comprising our cryo-EM structures sampled a significantly greater conformational space than would be expected from analysis of the ensemble structures alone (Figure 3 - figure supplement 2). Most intriguingly, the conformational changes displayed an unexpectedly high degree of change in the D1-D2-D3-D4 COM dihedral angle, indicating a significant rotation of IDE-N relative to IDE-C (Fig. 3B,C, Figure 3 – figure supplement 2).

Integration of the multibody results across all states of IDE suggest two dominant components of structural variance. First is a translation-dominant conformational change wherein IDE-N swings toward or away from IDE-C as if mediated by a hinge formed by the interdomain linker region of IDE (Fig. 3D, Figure 3 - video 1). This conformational change closely resembles our understanding of the IDE open-closed transition inferred from analysis of the ensemble structures. The second component is a rotation-dominant conformational change wherein IDE-N rotates orthogonal to the plane of the dimer as if it were being screwed into or ground against IDE-C (Fig. 3E, Figure 3 – video 2). Importantly, these dominant components of structural variance were also observed when the particle populations were analyzed with cryoSPARC’s 3DVA implementation, albeit over a smaller magnitude of conformational change (Figure 3 – figure supplement 3). These dominant components also correlate well with the lowest frequency modes revealed by normal mode analysis (Figure 3 – videos 3, 4). Interestingly, while the presence of insulin was found to substantially influence the consensus reconstructions of IDE, we found that the presence or absence of insulin had no significant effect on the principal components of structural variance for IDE, consistent with the current understanding that enzyme conformational changes are “hardwired” into the structure (Bahar et al, 2007; Bahar & Rader, 2005).

R668 plays a key role in the IDE open/close transition in all-atom MD simulations

Analysis of the particle heterogeneity comprising cryo-EM structures provides excellent information about the conformational space sampled by the protein of interest. However, such analysis lacks a temporal component, and thus only implies motions. To overcome this barrier and examine how the conformation of IDE changes over time, we performed six replicate all-atom MD simulations with explicit solvent. Simulations were initialized from the O/pO state (PDB: 7RZG), with the missing loops modeled in to generate continuous peptide chains and run for one microsecond each. By starting with the O/pO structure, we investigated the dynamics of both the open and closed subunits simultaneously. We observed that the conformational space sampled by our simulations correlated well with the conformational space sampled by our cryo-EM particle population, as revealed by multibody analysis (Figure 4 – figure supplement 1). In 5 of our 6 simulations, the open subunit closed quickly, typically in <200 ns (Fig. 4A). Analysis of the open- close transition revealed two key findings. First, the open subunits did not close to a singular structure. We define closing as a subunit reaching a similar D1-D4 COM distance as the pO or pC states, but we observed distinct differences among the subunits indicative of IDE-N rotation relative to IDE-C. These differences are in line with our second key finding, that the conformational changes associated with the open-close transition did not follow the same general pathway among the simulations, as we observed a high degree of variability upon closing (Fig. 4B).

All-atom MD reveals a molecular basis for IDE conformational dynamics.

(A) Measurements of the O subunit D1-D4 distance over the course of six separate microsecond long all-atom MD simulations of WT IDE. Of which, the open subunit closed in 5 of the 6 simulations. (B) Plot of the O subunit D1-D4 distance vs the D1-D2-D3-D4 COM dihedral angle over the course of the simulation of WT IDE. The open subunits displayed a variety of closing pathways and did not close to a consensus structure. Starting structure shown as black dot, pO structure shown as white dot. (C) R668 acts as a guidepost residue, rapidly interacting with D309 or E381. Formation of this interaction is associated with rapid closing, as measured by a decrease in D1-D4 distance (D). (E) Hydrogen-deuterium exchange mass spectrometry highlights the importance of R668 in mediating the open-close transition. In the presence of insulin (panel 1, red), Aβ (panel 2, red), and BDM-44768 (panel 3, red), all of which promote IDE closing, the peptide containing R668 shows reduced deuterium exchange relative to apo-IDE (black), yet in the presence of 6bk (panel 4, red), which does not promote closing, there is no difference in the exchange rates for the R668 containing peptide relative to apo-IDE (black). Helix containing R668 colored by red (increase) – white (no change) – blue (decrease) gradient depicting the degree of deuterium exchange relative to apo-IDE. (F) Measurements of the O subunit D1-D4 distance over the course of six separate microsecond long all-atom MD simulations of IDE R668A. (G) Plot of the O subunit D1-D4 distance vs the D1-D2-D3-D4 COM dihedral angle over the course of the simulation of IDE R668A. The six separate microsecond long simulations indicate that an R668A mutation significantly alters the closing dynamics of IDE (F) and increases the rotational motion (G) relative to WT (panels A and B respectively). Starting structure shown as black dot, pO structure shown as white dot.

Noticeably, R668 consistently stood out as one of the first IDE-C residues to form new interactions with IDE-N, either D309 or E381 despite the path variation among our MD simulations (Fig. 4C). In most simulations, formation of this interaction preceded a large decrease in the D1-D4 COM distance (Fig. 4D, Figure 4 – figure supplement 2). Interestingly, these R668 interactions are not present in the ensemble cryo-EM or crystal structures of IDE. We hypothesize that the biochemical properties and spatial localization of R668 enable it to essentially reach out and grab onto the N-domain, initiating the formation of a complex hydrogen bonding network that drives the open-close transition towards completion. Previously published hydrogen-deuterium exchange mass spectrometry (HDX-MS) data indicates that R668 is stabilized under conditions that promote IDE closing, suggesting that R668 may function as a molecular latch (Fig. 4E) (Zhang et al., 2018).

To further probe the importance of R668, we ran 6 additional simulations with IDE carrying a R668 to alanine (R668A) point mutation under identical conditions as described previously. These simulations displayed significantly altered dynamics when compared to the wild-type simulations. Notably, the R668A mutant was found to close more slowly, and many simulations did not reach the D1-D4 COM distance typified by a closed state structure, instead stabilizing with a significantly larger D1-D4 COM distance than we had previously observed in our wild type (WT) simulations or our experimental structures (Fig. 4F). We also observed that the R668A mutant preferentially sampled a subset of conformational space rarely explored in our WT simulations and displayed significantly greater variation in the D1-D2-D3-D4 dihedral angle (Fig. 4G). As a result of this increased rotational exploration, several of our R668A simulations closed to a conformation where IDE-N is offset and rotated relative to IDE-C when compared to the experimentally determined structures (Figure 4 – figure supplement 3). This altered conformation results in a structure where IDE-N and IDE-C are fully engaged, but the offset produces a deceivingly high D1-D4 COM distance. With this in mind, we sought to better understand the conformational dynamics of IDE in the closed state.

R668A modulates IDE conformational dynamics in vitro

The R668A mutation was found to significantly alter the conformational dynamics of IDE’s open- close transition in our simulations (Fig. 4F, G), so we expressed and purified the R668A mutant to determine if the changes predicted from our MD simulations resulted in altered enzymatic activity relative to WT. Previous work has established that IDE adopts a dominant O/pO state in solution (Zhang et al., 2018). Consistent with our MD simulations, size-exclusion chromatography revealed that the R668A construct eluted slightly earlier than WT IDE, suggesting that the R668A mutation induces a greater proportion of molecules to adopt a more open conformation, possibly shifting to a dominant O/O state (Fig. 5A). Next, we compared the enzymatic activity of the R668A and WT constructs using the fluorogenic substrate (7- methoxycourmarin-4-yl)acetyl-RPPGFSAFK(2,4-dinitrophenyl)-OH (substrate V), a bradykinin mimetic which has previously been used to characterize the enzymatic activity of IDE and related enzymes (King et al, 2014; McCord et al., 2013). We found that WT IDE has activity consistent with previously published data (McCord et al., 2013) and the R668A mutation produced a ∼5-fold decrease in activity (Fig. 5B). ATP is known to promote IDE degradation of small substrates, i.e. bradykinin, but not larger substrates, i.e. insulin (33). Consistent with this observation, the addition of ATP increased the activity of both WT and R668A constructs (Fig. 5B). We also tested if the R668A mutation would alter the degradation of larger, well-folded substrates, using insulin in a direct competition assay. We found that, after accounting for the previously observed 5x decrease in activity, insulin yielded an apparent Ki of ∼8 nM for WT IDE, but this value was reduced to ∼52 nM for the R668A construct (Fig. 5C). Conversely, when Aβ(1-40), an unstructured substrate, was tested in the same competition assay, we observed an apparent Ki of ∼688 nM for WT IDE, yet the R668A mutant did not exhibit a dose-dependent response to the presence of Aβ and the data could not be fit to any kinetic equation (Fig. 5D). We then performed size exclusion chromatography-coupled small-angle X-ray scattering (SEC-SAXS) for both constructs to ascertain if the observed differences in enzymatic activity could be explained by altered biophysical properties. The WT scattering profile produced an Rg of 49.4 +/- 0.9 Å, consistent with previous SEC-SAXS results (Fig. 5E, F) (Zhang et al., 2018). The R668A scattering profile produced an Rg of 54.7 +/- 0.4 Å, consistent with our SEC data that R668A mutant has slightly larger hydrodynamic radius (Fig. 5A). Together, our data indicates that R668A mutation can profoundly affect the conformational dynamics governing the open-closed transition, and that altered conformational dynamics can alter enzymatic activity in a substrate-dependent manner.

R668A alters IDE activity in vitro

(A) Elution profile of WT IDE (blue) compared to the R668A mutant (orange) from a S200 SEC column. (B) Degradation of the fluorescent substrate MCA-RPPGFSAFK(Dnp) by WT IDE and the R668A construct in the presence and absence of ATP. Data represents the average initial velocities of three replicates performed at a protein concentration of 3.125 nM. Error bars (gray) represent the standard error. (C) Inhibition of MCA-RPPGFSAFK(Dnp) degradation by WT IDE (circles, solid fit lines) and IDE R668A (squares, dashed fit lines) in the presence of varying amounts of insulin. Data was fit to the Michaelis-Menten (black) and Hill equations (red). Relevant parameters, Michaelis-Menten: WT: χ2=0.001, Vmax=0.951, Ki=8.3 nM; R668A: χ2=0.005, Vmax=0.892, Ki=52 nM; Hill: WT: χ2=0.009, n=0.55, Ki=51 nM; R668A: χ2=0.055, n=0.61, Ki=198 nM. Error bars represent standard error, data points represent the average of three replicates. (D) Inhibition of MCA-RPPGFSAFK(Dnp) degradation by WT IDE (circles, solid fit lines) and IDE R668A (squares, dashed fit lines) in the presence of varying amounts of insulin. WT data was fit to the Michaelis-Menten (black) and Hill equations (red), R668A data could not be fit to either equation, indicating that the mutation confers substrate-specific altered enzyme kinetics. Relevant WT parameters, Michaelis-Menten: χ2=0.066, Vmax=0.823, Ki=353 nM; Hill: χ2=0.08, n=0.72, Ki=688 nM. (E) SEC-SAXS profile of WT (black) and R668A (red) constructs with Rg values calculated by both the Guinier and Porod methods along with Dmax derived from the P(r) function (F).

Charge-swapping at the IDE-N/C interface mediates conformational dynamics in the closed state

The WT pO subunit remained closed throughout all simulations yet was found to initially relax to a slightly more open state within the first 200 ns, roughly the same time frame it took the WT O subunit to close (Fig. 6A) and, despite not opening, demonstrated a range of D1-D2-D3-D4 dihedral variation similar to that of the O subunits (Fig. 6B). Interestingly, the conformational variation within our simulations increased greatly after this relaxation point was reached (Fig. 6A). This relaxed conformational state consistently displayed an altered IDE-N/C interface compared to the experimentally determined structures that was replicated by the open WT subunits upon closing, leading to a nearly symmetric pO/pO state (Figure 6 – figure supplement 1). Biochemically, this relaxed IDE-N/C interface makes more sense than the interface observed experimentally. It has previously been demonstrated that the experimental constraints of crystallographic structure determination force IDE to adopt a closed conformation and determination of the open state can only be accomplished once those restraints are removed, i.e. cryo-EM (Zhang et al., 2018). These results suggest that both ensemble methods of structure determination may impact the conformation of IDE, however slightly. The IDE-N/C interface is littered with charged residues, yet few interactions are observed in the ensemble structures (Fig. 6C), whereas our MD simulations reveal a complex hydrogen bonding network (Fig. 6D). We observed an extensive network at the D2-D3 interface, with sporadic patches of interactions between D1 and D4 (Figure 6 – figure supplement 2). Within the D2-D3 network R668 again stands out as a notably key residue. In addition to forming hydrogen bonds with D309 and E381, we observed R668 form π-π interactions with R311 (Fig. 6E). Arginine-mediated π-interactions have been well characterized in other systems for their ability to stabilize interaction interfaces and facilitate conformational changes (Armstrong et al, 2016; Vernon et al, 2018).

Structural basis of closed state conformational dynamics.

(A) Measurements of the pO subunit D1-D4 distance over the course of six separate microsecond long all-atom MD simulations of WT IDE. (B) Plot of the pO D1-D4 distance vs the D1-D2-D3-D4 COM dihedral angle over the course of the simulation of WT IDE. (C) IDE-N/C interface previously solved crystal structures (PDB:2G47 shown) shows side chains are ill-positioned for interaction. (D) IDE-N/C interface formed upon open subunit closing in our MD simulations reveals a complex hydrogen bonding network. (E) Heat map showing conformational geometries that were preferentially sampled in our MD simulations by the open subunits upon closing. Insets highlight how the IDE-N/C interface changes to permit interdomain motion. (F) Plot of the O subunit D1-D2-D3-D4 dihedral angle during a subset of a single WT IDE MD simulation after the open-close transition has been completed. Charge-swapping between residues at the IDE-N/C interface is associated with changes in the D1-D2-D3-D4 dihedral. (G) For most of the simulation, D309 interacts with K483 (black), however, this interaction is broken for ∼100 ns, during which D309 instead interacts with R311 (blue) and R668 (orange). (H) For most of the simulation, D426 interacts with K571 (black), yet this interaction is periodically broken, and D426 instead interacts with K425 (green) and K899 (magenta). When these events of charge-swapping coincide with D309 charge-swapping (G), they are associated with a large change in the D1-D2-D3-D4 dihedral angle (F). When they occur alone, the effect on D1-D2-D3-D4 dihedral is smaller.

The IDE-N/C interactions we observe in our simulations are not restricted to discrete cognate pairs; rather, we observed that residues within the network periodically swap among several interaction partners. This “charge-swapping” phenomenon allows IDE-N to rotate or slide against IDE-C to adopt multiple conformations and maintain favorable interacting contacts (Fig. 6F). While multiple, concurrent, instances of charge-swapping were associated with rapid changes in the D1-D2-D3-D4 dihedral angle, singular events of charge-swapping did not appear to be significant enough to stimulate large-scale conformational change (Fig. 6G, H). Instead, substantial conformational changes appear to be caused by multiple charge- swapping events occurring in conjunction. This may explain why we observed no noticeable change in the conformational dynamics of the pO state subunits when comparing the WT and R668A constructs. Consurf analysis reveals that most of the residues forming these interaction networks are highly conserved among IDE homologs (Figure 6 – figure supplement 3) (Ben Chorin et al, 2020). This indicates that the revealed rotational motion is likely evolutionarily conserved and offers a potential mechanism by which IDE unfolds and repositions bound peptide substrates to degrade amyloid peptides.

Time-resolved cryo-EM suggests insulin-induced allostery in IDE

To begin understanding of how the interaction between IDE and insulin leads to insulin unfolding and degradation, we first performed all-atom MD simulation of insulin-bound IDE in the closed state. Unfortunately, we did not observe a noticeable structural change of insulin within our microsecond simulations. Previously, we used time-resolved SAXS analysis to determine the time constant for the insulin induced open to closed transition, tau (1, a.k.a., monoexponentially decay) to be ∼100 milliseconds (Zhang et al., 2018). However, it has previously been determined that IDE cleaves about 2 insulin molecules every second (Manolopoulou et al., 2009). We employed spotiton to rapidly mix and freeze grids where IDE has been exposed to insulin for 123 milliseconds prior to freezing (Dandey et al, 2020). We hoped that such an experimental setup would permit the identification of novel insulin-bound states of IDE and offer insight into the initial binding and unfolding of insulin. Previous work has established that, in the absence of substrate, IDE adopts a dominant O/pO state in solution and the presence of a molar excess of substrate shifts the particle population to a dominant pC/pC state (Zhang et al., 2018). Surprisingly, after processing the dataset in Relion, we observed a single conformational state within our rapidly-mixed dataset: the O/O state (Fig. 7A, Figure 7 - figure supplement 1, supplementary file 2). We were unable to resolve the structure beyond moderate resolution (∼5.1 Å) and did not observe unambiguous density indicative of insulin binding within the catalytic chamber, which is unsurprising given our previous observations of promiscuous binding (Fig. 2C). Despite the relatively high particle number in our final reconstruction, we were unable to sub-classify distinctly different conformations of IDE within the particle population as initial classes converged back to the dominant O/O state upon subsequent refinement. This leads us to believe that the obtained O/O state represents the average state of a particle population that exhibits a high degree of continuous structural heterogeneity. The O/O state observed in our rapidly-mixed dataset is distinct from the previously observed O/O states with D1-D4 center of mass distances of 55.1 Å and 46.8 Å matching the largest and smallest, respectively, distances previously observed for individual O state subunits (Fig. 7B). The discrepancy between the “openness” of the two subunits nearly doubles that observed in previous structures, making this O/O state the most asymmetric O/O state yet observed (Fig. 7C). The observation of a dominant O/O state in the presence of insulin under short timescales suggests that the binding of insulin to the O subunit of the apo-dominant O/pO state can induce the pO state to open through an unknown allosteric mechanism to reach the observed dominant O/O state (Fig. 7D). Positive allostery in IDE has previously been identified in biochemical assays, yet we lack a detailed understanding of the underlying mechanism (McCord et al., 2013; Song et al, 2011).

Time-resolved cryo-EM of IDE+insulin reveals a new O/O state.

(A) 5.1 Å reconstruction of IDE+Fab rapidly mixed with a 5x molar excess of insulin and vitrified with a mix-to-freeze time of 123 milliseconds. (B) Measurements (as in Fig. 2) of the distance between the D1-D4 centers-of-mass and dihedral angle formed by the D1-D2-D3-D4 centers-of-mass as indicators of the “openness” of the structure depicted in (A). (C) Alignment of the O/O state from our rapid-mixing dataset (domains colored as in Fig. 1) compared to the previously solved O/O states (gray, as labeled). (D) Overview of the conformational states adopted by IDE in all 4 available cryo-EM datasets with corresponding percentage of particles mapping to each state (Zhang et al., 2018)(this work).

Coarse-grained MD simulations reveal intermolecular cross-β sheet formation between insulin and IDE in all four IDE domains

As cryo-EM proved unsuitable to explore the mechanism of insulin unfolding by IDE, we turned to MD simulations. As stated above, there is a notable discrepancy in the time frame between the open/close transition and insulin cleavage rate. This suggests that unfolding requires multiple cycles of the open-closed transition and, thus it is not currently realistic to use all-atom MD simulations to probe the unfolding of insulin by IDE. We then used a coarse-grained MD package, Upside, to probe the interaction between IDE and insulin. Upside performs the dynamics simulation for just 6 atoms per residue with the sidechains being represented by directional-dependent beads whose packing probabilities have the lowest free energy while still including key structure details to compute realistic forces (Faruk et al, 2022; Faruk et al, 2023; Jumper et al, 2018a, b). This allows rapid equilibration and sampling of the energy landscape when compared with the conventional all-atom MD simulation, achieving ∼104 acceleration for the calculation time. Consistent with all-atom MD simulations, Upside simulations revealed the wide range of motions, including grid and hinge motions found in our all-atom MD simulation (Figure 8 – figure supplement 1). We observed the partial unfolding of insulin in our Upside simulations yet, importantly, the unfolding did not disrupt the inter- or intra-molecular disulfide bonds (Figure 8 – video 1).

Upside analysis of IDE-insulin interactions.

(A) Transient cross-β-strand interactions between insulin and IDE. Secondary structure (gray: coil, green: β-strand, yellow: α-helix) for each insulin residue (Y-axis) over the course (X-axis) of a representative Upside simulation. Points indicate predicted hydrogen bond interactions between IDE and insulin, colored by which domain of IDE that insulin residue is interacting with (D1: red, D2: orange, D3: blue, D4: cyan). Stars below the X-axis indicate frames from which structures were extracted for panel B. (B) Structure of IDE/insulin complex in Upside simulations at frames indicated in (A). IDE domains colored as in (A), insulin shown in magenta (chain A) and yellow (chain B). Inset highlights the intermolecular cross-β sheets formed between IDE and insulin within each of the four homologous IDE domains. (C) Individual domains of IDE, colored as in (A), extracted from (B) and oriented to demonstrate the structural conservation of intermolecular cross-β-sheet formation with bound insulin (colored as in B).

As insulin partially unfolds, it readily adopts β-strand structures—a key characteristic of amyloidogenic peptides—which can transiently form cross-β-sheets with the central β-sheet present in all four domains of IDE (Figure 8A-B). Specifically, we observed that cross-β-sheet interactions occur at distinct regions of insulin: the N-terminus of insulin chain A interacts with the IDE D2 domain near the substrate N-terminal anchoring site (also known as the exosite); the middle portion of insulin chain A interacts with the IDE D1 domain; and the C-terminal regions of insulin chain B interact with the D3 and D4 domains of IDE. While the interactions of insulin with IDE-C domain are novel, those with IDE D1/D2 domains are structurally similar to those previously observed (Guo et al, 2010; Manolopoulou et al., 2009; Ren et al, 2010; Shen et al., 2006). The labile and transient β-strand-rich interactions have been shown to mediate diverse protein-protein interactions, e.g., the assembly of intermediate filaments (Eibauer et al, 2024; Zhou et al, 2021).

This interaction provides molecular insight into how IDE recognizes and binds amyloidogenic peptides with diverse sequences (Figure 8C). IDE is capable of forming cross-β interactions with many regions of insulin, primarily mediated by backbone-to-backbone contacts that do not rely on specific amino acid sequences. The structural arrangement of IDE’s four conserved domains enables it to interact with multiple regions of its substrates, such as insulin, which are predisposed to forming β-strands (Figure 8C). We term this type of interaction "β-grabbing," and propose that it is a key mechanism by which IDE unfolds insulin. This interaction effectively tethers the substrate to both the IDE-N and IDE-C domains, facilitating the interdomain motions of IDE that promote substrate unfolding.

Discussion

Our results allow us to put forth a refined model describing how IDE recognizes amyloid peptides that are diverse in size and shape (Fig. 9). The catalytic cycle of IDE starts with at least one subunit of dimeric IDE adopting the open state, exposing the interior of the catalytic chamber to potential substrates (Zhang et al., 2018). The unconstrained motion of IDE-N likely facilitates substrate interaction in a variety of initial orientations. IDE-N and IDE-C have negatively and positively charged surfaces, respectively, and are able to attract peptide substrates with complementary charge profiles (Shen et al., 2006; Zhang et al., 2018). Such peptides have a high dipole moment and are often prone to aggregation. However, in the open state, the door subdomain, which contains the key catalytic residues, is highly flexible, rendering the open state catalytically incompetent (McCord et al., 2013; Zhang et al., 2018). Proper positioning and unfolding of substrate is necessary to stabilize the cleft prior to catalysis. Substrate binding has been suggested to enhance IDE closing through charge complementarity (Shen et al., 2006). It is likely that the rotation of IDE- N permits a range of closing pathways to accommodate non-optimal substrate binding geometry (Fig. 4). If the closing geometry permits R668 to interact with E381 or D309, the closing reaction continues to completion. If the closing geometry does not permit the R668-mediated interactions, this could be a signal of improper substrate binding, allowing IDE-N to open and either release improper substrate or reposition for another closing attempt. Once closed, IDE-N is capable of rotating against IDE-C mediated by charge- swapping events at the IDE-N/C interface (Fig. 6). The combined motions associated with the open to closed transition and rotation between IDE-N and IDE-C can lead to the distortion of the secondary structure of bound substrate. Furthermore, the transient interaction of insulin with the catalytic chamber of IDE such as β-grabbing via the β-strand pairing should further stress the integrity of bound substrate (Fig. 8). Together, these forces should facilitate unfolding and proper positioning of substrate for cleavage. Our data indicates that IDE is able to accommodate any possible orientation of insulin tethering within the catalytic site (Fig. 2C), suggesting that IDE is able to cut both chains of insulin in rapid succession without requiring significant repositioning (Manolopoulou et al., 2009). IDE then opens to release cleavage products and the cycle begins anew.

Model for the catalytic cycle of IDE.

The details are described in discussion. For simplicity, only the scenario involving the addition of a second insulin molecule is illustrated, although this is not required for the catalytic cycle. The mechanism by which IDE transitions between closed and open states remains unknown and is therefore indicated with question marks.

IDE utilizes both unfoldase and protease activities to degrade clinically-relevant peptides, including three glucose-regulating hormones with contrary effects – insulin, amylin, and glucagon – and Aβ, the accumulation of which is associated with the progression of Alzheimer’s disease. This has made IDE an attractive target for therapeutic intervention (Maianti et al., 2019; Tang, 2016). One major challenge is how to better control selectivity when the substrates IDE degrades are highly diverse in sequence and structure. It remains unknown how IDE unfolds substrates prior to cleavage. Our studies characterize the conformational dynamics of IDE and offer insight into the catalytic mechanisms, revealing that IDE-N rotates against IDE-C. We envision that this motion works in concert with the open-close transition to reposition and unfold substrates prior to cleavage. This rotation appears to be largely unconstrained in the open state and may play a role in promoting substrate capture. In the closed state, we found the rotational motion to be mediated by charge-swapping events at the IDE-N/C interface. These events are supplemented by interactions between hydrophobic patches that provide a non-specific interaction interface and allow the domains to slide against one another. We found that IDE-N/C interactions were mediated by the key residue

R668, and mutation of this residue altered the enzymatic activity and conformation of IDE, likely by impairing the open-close transition. Our data suggests that, while the R668A mutation globally impaired IDE activity, R668A exhibits significantly different kinetic profiles towards well-folded substrates (insulin) compared to unstructured substrates (Aβ). This highlights the complexity of the IDE catalytic cycle as we envision that IDE conformational dynamics can regulate solvent access to the catalytic chamber and substrate binding, unfolding, and cleavage are all intimately linked.

Our work adds to the expanding body of literature that integrates computational and experimental approaches (Arenz et al, 2016; Hirschi et al, 2021; Jiang & Doudna, 2017; Palermo et al, 2016; Roh et al, 2020; Saha et al, 2022). By employing multibody analysis, we identified the dominant sources of structural heterogeneity within our cryo-EM particle population, uncovering that the conformational dynamics of IDE involve a substantial rotational component that is not evident from ensemble structure analysis alone. We showed that all-atom MD simulations can address limitations of cryo-EM single particle analysis when investigating the role of residue R668. Because R668 forms a network of polar and electrostatic interactions rather than exclusive partnerships, dynamic hydrogen bonding networks at the IDE-N/C interface— particularly those involving R668—contribute to conformational heterogeneity, which further restricts the achievable resolution in cryo-EM analysis. Furthermore, vitrification by freezing proteins in liquid ethane can significantly alter the entropic contribution of R668 through conformation-dependent promiscuous contacts. By combining cryo-EM and MD simulations, we gained insight into how flexible interactions between IDE-N and IDE-C facilitate domain movements and highlighted the critical role of R668 in mediating these interactions. We also used MD simulations to demonstrate altered open-closed dynamics in the R668A mutant compared to the wild-type enzyme and validated these findings with experimental enzymatic kinetic assays and SEC-SAXS experiments.

At its core, cryo-EM is an ensemble method of structure determination, averaging tens of thousands of particles together to generate consensus structures. While methodologies have been well-established to separate out discrete conformational classes, proteins that exhibit constant gradients of structural heterogeneity remain problematic, although significant attention has been devoted to the issue in recent years, with varying levels of success (Nakane et al., 2018; Punjani & Fleet, 2021; Zhong et al, 2021). Currently, the best approaches use various dimensionality reduction techniques to analyze the principal components of particle heterogeneity. While a substantial innovation, it remains to be seen how the principal components of structural heterogeneity, derived from a series of static snapshots of a particle at the time of freezing, correlate with the real molecular motions of a protein in an aqueous environment. For many systems, especially enzymes of clinical significance, proper characterization of the conformational dynamics and transitional states of the protein are necessary for a full understanding of protein function and lays the foundation for future therapeutic development. MD simulations can provide this information, yet such studies are often met with skepticism by experimental researchers. By integrating MD simulations with experimental techniques, the weaknesses of one technique can be compensated for by the strengths of other techniques to provide comprehensive information about the system of interest. While this integrative approach has proven beneficial across a wide range of systems, including proteorhodopsin, the ribosome, and rotary ATPases, perhaps the best example of this approach is the CRISPR-Cas9 system, where an exhaustive number of studies have combined cryo-EM, crystallography, and traditional biophysical techniques with traditional and accelerated MD methods to drive the field forward at an astounding pace (Arenz et al., 2016; Hirschi et al., 2021; Jiang & Doudna, 2017; Palermo et al., 2016; Roh et al., 2020; Saha et al., 2022).

Our integrative structural analyses provide the structural basis for IDE conformational dynamics and their roles in IDE catalytic cycle and the framework for future studies. We speculate that multi-cycles of open-close transition and rotation driven by IDE conformational dynamics will be required for its unfoldase activity. However, the kinetic timescale of this reaction is fast, estimated to be ∼500 milliseconds rendering investigation challenging experimentally while the timescale is long for all-atom MD simulation (Manolopoulou et al., 2009). Our cryo-EM analyses also shed light on IDE allostery (McCord et al., 2013; Song et al., 2011; Zhang et al., 2018) (Fig. 9). The IDE dimer primarily adopts the O/pO conformational state (Zhang et al., 2018). Through time-resolved cryo-EM analysis, our data suggest that, upon substrate binding to the O state, the substrate-inaccessible pO state undergoes allosteric opening, allowing this subunit to capture its substrate. Additionally, the prolonged incubation of IDE with sub-stoichiometric amounts of insulin results in the predominance of the pC/pC state, suggestive of coordination in substrate capture and open-closed transition. Collectively, these positive allosteric transitions would enhance IDE’s capacity to capture and degrade its substrates more efficiently, though the underlying mechanisms remain unclear. Future integrative structural studies are required to further elucidate and control the IDE catalytic cycle, enabling substrate-specific modulation of IDE activity and advancing its therapeutic potential.

Materials and Methods

Expression and purification of human IDE

Cysteine-free IDE (referred to as IDE in this study) was expressed in E. coli BL21 (DE3) cells (at 25 °C for 20 h, 0.5mM IPTG induction using T7 medium). The R668A mutation was introduced into this construct by site-directed mutagenesis with the following primer pair: CCGGAAATTGTTAAGAGATGCCATATATGCTTCTTTG (forward) CAAAGAAGCATATATGGCATCTCTTAACAACCGGGC (reverse) and verified by sequencing. Recombinant IDE proteins were purified by Ni-NTA, source-Q, and Superdex 200 columns as previously described (Zhang et al., 2018). Protein was aliquoted and stored at -80° C.

Cryo-EM data collection and analysis

Thawed IDE aliquots were further purified by Superdex 200 chromatography using buffer containing 20 mM HEPES, pH 7.2, 150 mM NaCl, 10mM EDTA and then mixed with FabH11-E at an equal molar ratio. FabH11-E-IDE complex was purified by Superdex 200 chromatography in the presence of insulin at molar ratio IDE:insulin=2:1. Insulin was purchased from SIGMA (91077C). All grids, 300 mesh carbon or gold holey nanowire grids were plasma cleaned with O2 and H2 for 10 secs using a Solarus plasma cleaner (Gatan) and plunged at 133 milli-second using Chameleon, Spotiton-based technology. All images were acquired using a Titan Krios microscope (FEI) operated at 300KeV with a Gatan K3 direct electron detector (Gatan) in counting mode. Images were automatically acquired using Leginon (Suloway et al, 2005). Images were processed using software integrated into RELION3 (Zivanov et al, 2018). Frames were aligned using MotionCor2 (Zheng et al, 2017) software with dose weighting, CTF was estimated using Gctf (Zhang, 2016), particles were picked and extracted automatically using RELION. Particle stacks were processed through several rounds of 2D and 3D classification. Selected classes were then processed for high-resolution 3D refinement. The flowchart and detailed data processing is summarized in Figure 2 – figure supplement 1. Finally, the overall map was improved by particle polishing in RELION and sharpening. The final resolution was estimated using Fourier Shell Correlation (FSC=0.143, gold-standard). The density fitting and structure refinement was done using UCSF CHIMERA (Pettersen et al, 2004), COOT (Emsley & Cowtan, 2004), REFMAC5 (Murshudov et al, 2011) and PHENIX (Adams et al, 2010). Time-resolved cryo- EM grids by rapidly mixing IDE+Fab with insulin were prepared using the Spotiton device as previously described (Budell et al, 2021; Dandey et al., 2020). In brief, IDE+fab and insulin were thawed on ice and diluted in 20 mM HEPES, pH 7.2, 150 mM NaCl, 10 mM EDTA to 25 µM and 127.5 µM, respectively. Self- wicking, nanowire grids were plasma treated (Gatan Solarus) at 5 W in hydrogen and oxygen for 1-2 minutes (Wei et al, 2018). Each sample was aspirated into separate piezodispensers and applied sequentially in 70-100% relative humidity at 22-25C onto plunging grids just before cryogen entry. The maximum constant velocity of the plunging robot was adjusted to obtain the target sample mixing times of 123 milliseconds. The data was collected and processed as above, in Relion5, with Blush regularization being employed during 3D refinement.

Multibody analysis

We defined three rigid bodies as shown in figure 2A for the analysis. The size of the user-defined bodies has been theorized to play a significant role in the success of multibody refinement, most likely due to the fact that larger bodies will have a stronger signal-to-noise ratio (Nakane et al., 2018). With a Fab bound to each of IDE-N within IDE dimer, we assessed the impact of body size on multibody refinement by examining how density corresponding to 2 Fv regions, 1 Fv region, or no Fv regions would affect the multibody refinement. Fab has a variable region (Fv), and a constant region (Fc) and we chose to subtract and mask out the density corresponding to the Fab Fc, as we speculated that any motion between the Fab Fv and Fc domains would bias the results of the multibody refinement. The quality of the multibody results was assessed based on the subsequent improvement in the Coulomb potential map quality and calculated resolution. In the absence of any Fv density, the maps resulting from multibody refinement had the best calculated resolution, yet the map quality was quite poor overall; much of the density appears globular and featureless, particularly in the IDE-N regions (Figure 2 – figure supplement 4). Conversely, when Fv density was present on both IDE-N bodies, the resolution of the multibody output maps (4.5 Å) was worse than the resolution of the input map (4.3 Å). Thus, we found that the greatest improvement occurred when the Fv density was present on only the exterior body of the pO subunit. In this case, multibody refinement improved both the calculated resolution and density quality.

Following multibody refinement, the structural heterogeneity for each body was analyzed along six principal component vectors, resulting in 18 vectors describing the primary components of structural variation within the IDE dimer. We analyzed the variation along these vectors as a proxy for protein motion. The results of the multibody analysis indicate the degree of structural variation that is explained by motion along each of the primary component vectors. For all IDE states, we observed no single component vector could explain the majority of structural variance (Supplementary files 3-8). Given the underlying assumptions of multibody refinement, there is a question of whether any observed motions are biologically relevant or simply the result of improper particle alignment. We reasoned that the greater the degree of variation explained by motion along a specified component vector, the greater the likelihood was that the motion would be biologically relevant. Therefore, we focus on the top 9 component vectors, which cover ∼75% of total variance for each state, in greater detail.

Molecular dynamics simulations

All-atom MD simulations were prepared from the O/pO structure of IDE (PDB: 7RZK) and necessary CHARMM force field PARAM36 files using QwikMD (Ribeiro et al, 2016). Gaps in the structure were modeled in using the AlphaFold model of IDE. A three-step minimization-annealing-equilibration process was used to generate an equilibrated system. Simulations were performed under NPT (constant numbers of particles N, pressure P, and temperature T) conditions at 310 K and 1 atm with periodic boundary conditions in NAMD3.0 (Phillips et al, 2005). Explicit solvent was described with the TIP3P model (Jorgensen et al, 1983). Custom TCL scripts in VMD were used to calculate the D1-D4 centers-of-mass distance and D1-D2-D3-D4 centers-of-mass dihedral angle (Supplementary file 9). Residues 964-988 in D4 that are mostly absence in all static structures of IDE were built using Alpha-fold, which is highly flexible in our MD simulations. They were omitted from the analysis of MD simulations, as the inclusion of this flexible loop significantly altered the center-of-mass independent of global conformational change.

Initial structure of insulin bound IDE, for Upside simulations, was generated using PYMOL that started with IDE closed monomer (PDB code: 6BF8). Intact insulin (PDB code: 4INS) was fixed by mutation of residue 30 of insulin chain B back to threonine and modeled into the catalytic chamber based on the crystal structure of IDE in complex with the partially unfolded insulin (PDB code: 2WBY). The upside simulation was run with 28 replicates at TUpside=0.8 using Upside 1.0 force field parameters and our total upside simulation time is approximately 28 μsec (Faruk et al., 2022; Faruk et al., 2023; Jumper et al., 2018a, b). We estimated Upside’s time step to be ∼0.1 picosecond based on folding rates and chain motions. Custom R, and Python scripts were used for data analysis and presentation (Supplementary files 10-12).

Enzymatic assays

IDE constructs were exchanged into activity buffer comprised of 25 mM Tris-HCl pH 7.5, 150 mM NaCl, and 10 mM ZnCl2. MCA-RPPGFSAFK(Dnp) was purchased from Enzo Life Sciences (BMLP2270001), resuspended in dimethyl sulfoxide at a concentration of 5 mM and diluted to 5 μM with activity buffer. For activity measurements, 5 μM MCA-RPPGFSAFK(Dnp) was mixed with the desired IDE construct at concentrations ranging from 1-100 nM in the presence or absence of 1 mM ATP in a 200 μl reaction. Fluorescence was monitored every 30 sec for 30 minutes with excitation/emission wavelengths of 320/405 nm at 37° C. Initial velocity was calculated during the linear range. Insulin was purchased from MP Biomedicals (#0219390010), resuspended in 0.01N HCl at 1 mM and diluted to desired concentration with activity buffer. For competition assays, 5 nM IDE construct and 5 μM MCA-RPPGFSAFK(Dnp) was mixed with insulin ranging from 0-100 μM. Fluorescence was monitored every 30 sec for 30 minutes with excitation/emission wavelengths of 320/405 nm at 37° C. Initial velocity was calculated during the linear range for each concentration of insulin and normalized relative to the respective construct’s velocity in the absence of insulin to maintain consistency across plates, yielding values for relative inhibition which were plotted vs [insulin] or [Aβ] and fit to the Michaelis-Menten equation in xmgrace to generate apparent Ki values. All experiments were performed in triplicate.

Size-exclusion chromatography coupled small-angle X-ray scattering

SAXS/WAXS data collection was employed the Life Sciences X-ray Scattering (LIX) Beamline at the National Synchrotron Light Source II (NSLS II) at Brookhaven National Laboratory in Upton, NY. Briefly, 60ul of samples in solution were pipetted into PCR tubes, placed into a Bio-Inert Agilent 1260 Infinity II HPLC multisampler and measured using the isocratic SEC-SAXS format at the beamline (Yang et al, 2021). 50ul of sample was injected into the Phenomenex Biozen dSEC2 3 μm bead size with 200Å pore size column was utilized at a flow rate of 0.35mL/min for 25 minutes. SAXS and WAXS data are collected simultaneously on a Pilatus 1M (SAXS) and Pilatus 900K (WAXS) detectors with a 2 second exposure (Yang et al, 2020). Data from both detectors is then scaled and merged. Intensity is normalized using the water peak height at 2.0 Å -1. Data processing and analysis was performed using py4xs and lixtools in jupyter lab. Buffer frames 100-125 were averaged and used for subtraction of averaged frames under the peak of interest.

eLife Assessment

The manuscript by Mancl et al. provides important mechanistic insights into the conformational dynamics of Insulin Degrading Enzyme (IDE), a zinc metalloprotease involved in the clearance of amyloid peptides. In the revised version, the authors have substantially expanded their analysis by incorporating time-resolved cryo-EM and coarse-grained molecular dynamics simulations, which reveal an insulin-induced allosteric transition and transient β-sheet interactions underlying IDE's unfoldase activity. Supported by a convincing combination of cryo-EM, SEC-SAXS, enzymatic assays, and both all-atom and coarse-grained simulations, this work refines our understanding of IDE's functional cycle and offers a structural framework for developing substrate-selective modulators of M16 metalloproteases.

Reviewer #1 (Public review):

Summary:

Mancl et al. present a comprehensive integrative study combining cryo-EM, SAXS, enzymatic assays, and molecular dynamics (MD) simulations to characterize conformational dynamics of human insulin-degrading enzyme (IDE). In the revised manuscript, the study now also includes time-resolved cryo-EM and coarse-grained MD simulations, which strengthen the mechanistic model by revealing insulin-induced allostery and β-sheet interactions between IDE and insulin. Together, these results expand the original mechanistic insight and further validate R668 as a key residue governing the open-close transition and substrate-dependent activity modulation of IDE.

Strengths:

The authors have substantially expanded the experimental scope by adding time-resolved cryo-EM data and coarse-grained MD simulations, directly addressing requests for mechanistic depth and temporal insight. The integration of multiple resolution scales (cryo-EM heterogeneity analysis, all-atom and coarse-grained MD simulations, and biochemical validation) now provides a coherent description of the conformational transitions and allosteric regulation of IDE. The addition of Aβ degradation assays strengthens the claim that R668 modulates IDE function in a substrate-specific manner. Finally, the manuscript reads more clearly: figure organization, section headers, and inclusion of a new introductory figure make it accessible to a broader audience. Overall, the revision reinforces the conceptual advance that the dynamic interdomain motions of IDE underlie both its unfoldase and protease activities and identifies structural motifs that could be targeted pharmacologically.

Weaknesses:

While the authors acknowledge that future studies on additional IDE substrates (e.g., amylin and glucagon) are warranted, such experiments remain outside the present scope. Their absence modestly limits the generalization of the R668 mechanism across all IDE substrates. Despite improved discussion of kinetic timescales and enzyme-substrate interactions, experimental correlation between MD timescales and catalysis remains primarily inferential. The moderate local resolution of some cryo-EM states (notably O/pO) continues to limit atomic interpretation of the most flexible regions, though the authors address this carefully.

Reviewer #2 (Public review):

Summary:

The manuscript describes various conformational states and structural dynamics of the Insulin degrading enzyme (IDE), a zinc metalloprotease by nature. Both open and closed state structures of IDE have been previously solved using crystallography and cryo-EM which reveal a dimeric organization of IDE where each monomer is organized into N and C domains. C-domains form the interacting interface in the dimeric protein while the two N-domains are positioned on the outer sides of the core formed by C-domains. It remains elusive how the open state is converted into the closed state but it is generally accepted that it involves large-scale movement of N-domains relative to the C-domains. Authors here have used various complementary experimental techniques such as cryo-EM, SAXS, size-exclusion chromatography and enzymatic assays to characterize the structure and dynamics of IDE protein in the presence of substrate protein insulin whose density is captured in all the structures solved. The experimental structural data from cryo-EM suffered from high degree of intrinsic motion amongst the different domains and consequently, the resultant structures were moderately resolved at 3-4.1 Å resolution. Total five structures were generated in the originally submitted manuscript using cryo-EM. Another cryo-EM reconstruction (sixth) at 5.1Å resolution was mentioned after first revision which was obtained using time-resolved cryo-EM experiments. Authors have extensively used Molecular dynamics simulation to fish out important inter-subunit contacts which involves R668, E381, D309, etc residues. In summary, authors have explored the conformational dynamics of IDE protein using experimental approaches which are complimented and analyzed in atomic details by using MD simulation studies. The studies are meticulously conducted and lay ground for future exploration of protease structure-function relationship.

Comments after first peer-review:

The authors have addressed all my concerns, and have added new data and explanations in terms of time-resolved cryo-EM (Fig. 7) and upside simulations (Fig. 8) which in my opinion have strengthened the merit of the manuscript.

Author Response:

The following is the authors’ response to the previous reviews

Public Reviews:

Reviewer #1 (Public review):

Summary:

Mancl et al. present a comprehensive integrative study combining cryo-EM, SAXS, enzymatic assays, and molecular dynamics (MD) simulations to characterize conformational dynamics of human insulin-degrading enzyme (IDE). In the revised manuscript, the study now also includes time-resolved cryo-EM and coarse-grained MD simulations, which strengthen the mechanistic model by revealing insulin-induced allostery and β-sheet interactions between IDE and insulin. Together, these results expand the original mechanistic insight and further validate R668 as a key residue governing the open-close transition and substrate-dependent activity modulation of IDE.

Strengths:

The authors have substantially expanded the experimental scope by adding time-resolved cryo-EM data and coarse-grained MD simulations, directly addressing requests for mechanistic depth and temporal insight. The integration of multiple resolution scales (cryo-EM heterogeneity analysis, all-atom and coarse-grained MD simulations, and biochemical validation) now provides a coherent description of the conformational transitions and allosteric regulation of IDE. The addition of Aβ degradation assays strengthens the claim that R668 modulates IDE function in a substrate-specific manner. Finally, the manuscript reads more clearly: figure organization, section headers, and inclusion of a new introductory figure make it accessible to a broader audience. Overall, the revision reinforces the conceptual advance that the dynamic interdomain motions of IDE underlie both its unfoldase and protease activities and identifies structural motifs that could be targeted pharmacologically.

Weaknesses:

While the authors acknowledge that future studies on additional IDE substrates (e.g., amylin and glucagon) are warranted, such experiments remain outside the present scope. Their absence modestly limits the generalization of the R668 mechanism across all IDE substrates. Despite improved discussion of kinetic timescales and enzyme-substrate interactions, experimental correlation between MD timescales and catalysis remains primarily inferential. The moderate local resolution of some cryo-EM states (notably O/pO) continues to limit atomic interpretation of the most flexible regions, though the authors address this carefully.

Reviewer #2 (Public review):

Summary:

The manuscript describes various conformational states and structural dynamics of the Insulin degrading enzyme (IDE), a zinc metalloprotease by nature. Both open and closed state structures of IDE have been previously solved using crystallography and cryo-EM which reveal a dimeric organization of IDE where each monomer is organized into N and C domains. C-domains form the interacting interface in the dimeric protein while the two N-domains are positioned on the outer sides of the core formed by C-domains. It remains elusive how the open state is converted into the closed state but it is generally accepted that it involves large-scale movement of N-domains relative to the C-domains. Authors here have used various complementary experimental techniques such as cryo-EM, SAXS, size-exclusion chromatography and enzymatic assays to characterize the structure and dynamics of IDE protein in the presence of substrate protein insulin whose density is captured in all the structures solved. The experimental structural data from cryo-EM suffered from high degree of intrinsic motion amongst the different domains and consequently, the resultant structures were moderately resolved at 3-4.1 Å resolution. Total five structures were generated in the originally submitted manuscript using cryo-EM. Another cryo-EM reconstruction (sixth) at 5.1Å resolution was mentioned after first revision which was obtained using time-resolved cryo-EM experiments. Authors have extensively used Molecular dynamics simulation to fish out important inter-subunit contacts which involves R668, E381, D309, etc residues. In summary, authors have explored the conformational dynamics of IDE protein using experimental approaches which are complimented and analyzed in atomic details by using MD simulation studies. The studies are meticulously conducted and lay ground for future exploration of protease structure-function relationship.

Comments after first peer-review:

The authors have addressed all my concerns, and have added new data and explanations in terms of time-resolved cryo-EM (Fig. 7) and upside simulations (Fig. 8) which in my opinion have strengthened the merit of the manuscript.

We are grateful for the dedication and constructive feedback provided by the editors and reviewers. We have revised our manuscript according to the suggestions by both reviewers.

Recommendations for the authors:

Reviewer #1 (Recommendations for the authors):

The new version of the manuscript reads exceedingly well and the corrections the authors have made during their revision made the manuscript much easier to read and digest than the first version. Below are minor details that may be corrected:

Abstract:

Line 45-47: "IDE is known to transition between a closed state, poised for catalysis, and an open state, able to release cleavage products and bind a new substrate." (consider adding a)

Fixed

Line 48-50: "Combining cryo-EM heterogeneity analysis with all-atom molecular dynamics (MD) simulations, we identified the structural basis and key residues for IDE conformational dynamics that were not previously revealed by IDE static structures." (consider adding previously)

Changed

Line 52-54: "Our small-angle X-ray scattering analysis and enzymatic assays of an R668A mutant indicate a profound alteration of conformational dynamics and catalytic activity." (consider adding analysis)

Changed

Line 54: Consider leaving out "Upside" in the abstract (to avoid confusion when reading the abstract) and leave it to be introduced in the introduction when Upside MD simulations are first mentioned.

Changed

Results:

Figure 5D: There seems to be an error in the legend for Figure 5D. It says "... presence of varying amounts of insulin", but this must be Aβ1-40. Please add info on whether the replicates are technical or biological.

The legend has been revised as suggested.

Line 125: Consider switching the order of "here" and "we"

“here” has been removed.

Line 128: Replace "5" with "five"

Changed

Line 137: Replace "when insulin is present" with "in the presence of insulin"

Changed

Line 228: Replace "5" and "6" with "five " and "six"

Changed

Line 229: Consider adding the word "form": "First, the open subunits did not close to form a singular structure."

We have adjusted the sentence to read “close to a singular consensus structure”

Line 327: Replace "2" with "two"

Changed

Line 276: Consider replacing "Conversely" with a more suitable connecting term as it implies that the observation presented in the two sentences are reverse or rephrase what is being compared. Is it the fact there is a dose dependency or not between the substrates or is it the actual kinetic parameters that are described. I just don't think conversely is fair with the current formulation as "the R668A mutant did not exhibit a dose-dependent response to the presence of Aβ" not that the Ki is reduced for WT compared to the R668A construct when looking at Aβ.

The connecting term has been removed completely, beginning the sentence with “When Abeta…”

Line 359: Replace "6" with "six"

Changed

Consider getting rid of possessive apostrophes to keep a formal tone, e.g. lines 211 (cryoSPARC's), 259 (IDE's) and 382 (IDE's). Exception to this is Alzheimer's disease.

All instances of possessive apostrophes, aside from Alzheimer’s, have been replaced alter more formal wording.

Figure 7 supplement 1: The color scheme for the local resolution is missing the unit (Å).

This has been corrected.

Finally, the supplementary videos illustrating IDE conformational dynamics are difficult to interpret and somewhat redundant in their current form. The transitions occur very rapidly, making it hard to appreciate the described motions, and the uniform coloring of IDE further limits visual clarity. I apologize for not including this point in my initial review. I recommend either removing the videos or re-rendering them to improve interpretability, for example by slowing down the motion and applying the same domain color scheme introduced in the new Figure 1 (and used in the MD trajectory video). This would greatly aid readers in connecting the descriptions in the text to the visual representations in the movies.

Figure 3 videos 1-4 were slowed down, simplified, and recolored to improve clarity.

Reviewer #2 (Recommendations for the authors):

Comments after first revision for authors:

Thanks a ton to the authors for the detailed explanation on my comments. I believe the discussions will help a large group of audience, especially the non-experts. Please address the minor comment below:

Minor comment:

Please update Supplementary file 1 (Cryo-EM data collection, refinement, and validation statistics) regarding the new volume obtained by time-resolved cryo-EM. Kindly also check line 47 in the abstract: "Here, we present five cryo-EM structures" , which may need an update (six structures and resolution 3.0-5.1 Å) or rephrase the sentences accordingly. If similar instances are found in the manuscript, where list of all the structures are mentioned together, please update accordingly if necessary.

The cryo-EM statistics for the time-resolved cryo-EM are shown in supplementary file 2 to differentiated two datasets. The abstract has been changed, as has line 149.

Version 3

Abstract

Insulin degrading enzyme (IDE) is a dimeric M16A zinc metalloprotease that degrades amyloidogenic peptides diverse in shape and sequence, including insulin and amyloid-β, to prevent toxic amyloid fibril formation. IDE has a hollow catalytic chamber formed by two ∼55 kDa N- and C-domains (IDE-N and IDE-C, respectively), in which peptides bind, unfold, and are repositioned for proteolysis. IDE is known to transition between a closed state, poised for catalysis, and an open state, able to release cleavage products and bind a new substrate. Here, we present six cryo-EM structures of the IDE dimer at 3.0-5.1 Å resolution, obtained in the presence of a sub-saturating concentration of insulin. Combining cryo-EM heterogeneity analysis with all-atom molecular dynamics (MD) simulations, we identified the structural basis and key residues for IDE conformational dynamics that were not previously revealed by IDE static structures. Notably R668 serves as a molecular latch mediating the open-close transition and facilitates key protein motions through charge-swapping interactions at the IDE-N/C interface. Our small-angle X-ray scattering analysis and enzymatic assays of an R668A mutant indicate a profound alteration of conformational dynamics and catalytic activity. By integrating coarse-grained MD simulations, our analysis reveals that IDE unfolds its substrates through the coordinated motion between IDE-N and IDE-C, as well as β-sheet formation between IDE and insulin. Additionally, our time-resolved cryo-EM analysis uncovers IDE allostery within the IDE dimer. Collectively, our findings demonstrate the strength of combining experimental and computational approaches to probe protein dynamics and pave the way for developing substrate-specific modulators of IDE activity.

Introduction

Protein homeostasis (a.k.a. proteostasis) is maintained by three primary mechanisms: chaperones, ubiquitination/proteasome, and autophagy (Hipp et al, 2019; Lopez-Otin et al, 2013). Disruptions in proteostasis can lead to the accumulation of amyloid fibrils and subsequent human diseases (Chiti & Dobson, 2006; Chiti & Dobson, 2017; Greenwald & Riek, 2010). As a result, many proteases have evolved to specifically target amyloidogenic peptides (Malito et al, 2008; Saido & Leissring, 2012). Among those, the cryptidase family, which includes the M16 metalloproteases insulin degrading enzyme (IDE) use an internal catalytic chamber, or “crypt”, to capture and selectively degrade the monomeric form of amyloid peptides to control the formation of amyloid fibrils (Liang et al, 2022; Malito et al., 2008). IDE effectively degrades various bioactive peptides, including amyloid-β (Aβ), a peptide associated with the progress of Alzheimer’s disease, and three blood glucose-regulating hormones: insulin, amylin, and glucagon (Tang, 2016). Consequently, defects in IDE alter the progression of type 2 diabetes mellitus and Alzheimer’s disease in animal models and are linked to these diseases in humans (Bertram et al, 2000; Bertram et al, 2007; Farris et al, 2003; Farris et al, 2004; Kim et al, 2007; Sladek et al, 2007; Tang, 2016; Tanzi & Bertram, 2005; Zeggini et al, 2007). IDE is a promising therapeutic target, as its inhibition improves glucose tolerance, yet progress has been hampered by the contrary actions of its diverse substrate pool (Maianti et al, 2014; Maianti et al, 2019; Tang, 2016). For example, IDE overexpression has been shown to reduce Aβ loads, in mice, however, this reduction is accompanied by hyperglycemia from a similar reduction in insulin levels (Leissring et al, 2003).

IDE-mediated substrate recognition and degradation is a complex process. IDE is capable of degrading substrates with diverse sequences and structures, such as Aβ and the three glucose-regulating hormones: insulin, amylin, and glucagon (Malito et al., 2008; Tang, 2016). Despite this diversity, IDE exhibits high selectivity—for example, it preferentially degrades insulin and atrial natriuretic peptide over insulin-like growth factor-1 and brain natriuretic peptide, respectively (Malito et al., 2008). The presence of ATP, inositol phosphate, or specific mutations can markedly enhance the catalytic rate for small peptide substrates like bradykinin, but does not significantly affect the degradation of larger substrates such as insulin and Aβ (Im et al, 2007; McCord et al, 2013; Song et al, 2017; Song et al, 2004). Additionally, the binding location of IDE inhibitors within the catalytic chamber can influence substrate preference, making IDE more likely to degrade one substrate over another (Charton et al, 2014; Maianti et al., 2019). This complexity arises partly from the interplay between the unfoldase and protease activities of IDE, which are essential for degrading diverse substrates (Malito et al., 2008; Tang, 2016; Zhang et al, 2018). IDE relies on its unique large catalytic chamber and associated conformational dynamics to unfold and degrade amyloidogenic substrates. A deeper understanding of the molecular mechanisms by which IDE utilizes its catalytic chamber for substrate unfolding and degradation is crucial for effectively harnessing this enzyme in future therapeutic development.

IDE is a dimeric 110 kDa metalloprotease. Each protomer is comprised of four homologous subdomains organized into two ∼55 kDa domains, IDE-N and IDE-C, connected by a short flexible linker (Li et al, 2006; Shen et al, 2006) (Fig. 1A). Our previous cryo-EM analysis revealed the major conformational states of IDE. In the absence of substrate, each protomer was found to adopt an open (O) or partial open (pO) state which are primarily differentiated by the displacement of IDE-N relative to IDE-C, the combination of which led to three dimeric conformations (O/O, O/pO, pO/pO) (Zhang et al., 2018). It was observed that substrate binding induced both protomers of IDE to fully close, referred to as the partial closed (pC, pC/pC for the dimer) state, as the cryo-EM closed state was found to be slightly more open than previously solved crystal structures, likely influenced by the constraints of the crystal lattice (McCord et al., 2013; Shen et al., 2006). Only the O state permits substrate access to the catalytic chamber, yet the catalytic cleft is stabilized only in the pC state with substrate bound, requiring IDE to undergo a substantial open-close transition during catalysis, the specifics of which are unknown (Zhang et al., 2018).

Overview of structure and function of IDE.

(A) Overall structure of IDE. IDE is comprised of four structurally homologous domains: D1 (red, residues 43-285), D2 (orange, residues 286-515), D3 (blue, residues 545-768), and D4 (cyan, residues 769-1,016). The domains are arranged in two roughly hemispherical regions that enclose the catalytic chamber referred to as IDE-N (comprised of the D1 and D2 domains) and IDE-C (comprising the D3 and D4 domains), which are joined by a linker region (residues 516-544). (B) Primary sequence and overall structure of insulin. Insulin cleavage by IDE as revealed by mass spectrometry (Manolopoulou et al., 2009). Cleavage sites are marked in arrows, which occur within α-helical regions and requires the peptide to be unfolded prior to cleavage. (C) Overview of the IDE catalytic cycle with key questions addressed in this study.

Biochemical and structural analyses indicate that many IDE substrates, such as insulin, must first be unfolded prior to cleavage (Fig. 1B, C). The conformational dynamics of IDE are thought to play a central role in substrate recognition and degradation, yet several mechanistic details underlying the unfoldase activity of IDE remain unresolved (Fig. 1). We hypothesize that the unfoldase and protease activities of IDE are regulated by the relative motions of the IDE-N and IDE-C domains, their interactions with substrates, and allosteric communication between subunits within the IDE dimer. To investigate this, we solved cryo-EM structures of IDE bound to insulin at a 2:1 IDE:insulin ratio, as well as structures captured following rapid mixing of IDE with insulin under functionally relevant conditions. Utilizing newly developed computational methods, we analyzed particle heterogeneity in these structures to infer the conformational dynamics of IDE. By integrating these experimental data with all-atom and coarse-grained molecular dynamics (MD) simulations, we identified and modulated specific molecular interactions to manipulate key conformational dynamics of IDE, ultimately enhancing our understanding how to control its enzymatic activity in vitro.

Results

Cryo-EM reveals new IDE structures in the presence of a sub-saturating concentration of insulin

We have previously used cryo-EM to solve the structure of IDE in the presence of a 5-fold molar excess of insulin, resulting in a dominant pC/pC state with both substrate binding chambers yielding density corresponding to unfolded insulin (Zhang et al., 2018). To explore the mechanism of IDE-directed substrate unfolding under more physiologically relevant conditions, we investigated IDE-insulin interactions with an IDE:insulin molar ratio of 2:1. We collected a dataset of ∼7,600 micrographs on a Titan Krios operated at 300 keV, from which 7.2 million particles were picked for processing in RELION. Following an established workflow, and processing the data with C1 symmetry, we generated five structures: three previously observed structures with improved resolution (pC/pC, O/O, and O/pO at 3.0 Å, 3.8 Å, and 4.1 Å resolution) and two novel states (O/pC and pO/pC at 3.4 Å and 3.3 Å resolution, respectively) (Fig. 2A, Figure 2 – figure supplement 1, Supplementary file 1). Interestingly, the pC/pC was still found to be dominant, despite IDE and insulin being present at a 2:1 molar ratio, consistent with the allostery of IDE (McCord et al., 2013; Zhang et al., 2018).

Cryo-EM structures.

(A) Overview of the cryo-EM structures. See figure S1 for processing details. (B) Comparison of the open (O), partial open (pO), and partial closed (pC) subunit states present in our cryo-EM structures with domain organization. The distance between the D1 and D4 domain centers-of-mass (D1-D4 COM) along with the dihedral angle formed by the D1-D2-D3-D4 domain centers-of-mass (D1-D2-D3-D4 dihedral) described in Zhang et al. (21) and depicted in Figure 3 – figure supplement 1 were used as biologically important criteria to quantify observed conformations. (C) Insulin density and corresponding model in our cryo-EM structures. Both the A chain (magenta) and B chain (yellow) can fit the density in the exosite and catalytic cleft.

The individual protomers of IDE adopt the same three conformations: O, pO, and pC states, reported previously but there are significant differences in the O and pO states (Fig. 2B) (Zhang et al., 2018). While the pC/pC state is nearly identical to the two states reported previously (RMSD of ∼0.6 Å; PDB ID: 6BFC, 6B3Q) (Zhang et al., 2018), the O and pO states appear more closed in the presence of insulin, suggesting that the presence of insulin promotes the open to closed transition. Specifically, the pO and O subunits are 3-5 Å more closed based on the distance between the centers-of-mass (COM) of the IDE D1 and D4 subdomains, while the dihedral angle formed by the COM of the D1-D2-D3-D4 subdomains is reduced by 5-10° (Fig. 2B). Alignment of the new O/O and O/pO states to their previously solved counterparts reveals little difference, with a global RMSD of 1.316 Å (O/O, PDB ID: 6B7Y) and 1.405 Å (O/pO, PDB ID: 6BF8), respectively (Zhang et al., 2018). The primary source of variation between the different states remains the degree of opening between the two domains of each subunit, as a global alignment of IDE-N and IDE-C across all structures reveal the domains primarily act as rigid bodies (RMSD <1.4 Å). Consistent with previous observations, the door region, which contains the catalytic zinc binding site, exhibited higher B-factors in the O states than in either the pO or pC state (Zhang et al., 2018) (Figure 2 – figure supplement 2).

The cryo-EM structures presented here contain four subunits that have clear density present at the catalytic cleft and exosite indicative of bound substrate, thus adopting the pC state. Previous structures have modeled the same chain of insulin in both the exosite and catalytic cleft, but the best models we could build into our density suggest that the chain bound to the exosite is not the same chain positioned for cleavage at the catalytic cleft (Fig. 2C). The preponderance of data suggests it to be unlikely that insulin adopts a preferred orientation within the catalytic chamber, rather the catalytic chamber can accommodate insulin in four different orientations, with either the A or B chain bound to the exosite and catalytic cleft in cis and trans orientations. This promiscuity of binding has been theorized as a reason for why there is no chain preference for the initial insulin cleavage event (Manolopoulou et al, 2009). Crystal structures of insulin are highly compact and globular, with little spatial separation between the N-and C-termini of each chain, and the IDE cleavage sites are located within α-helices (Baker et al, 1988). In its crystallized state, insulin cannot interact with both the exosite and catalytic cleft unless it unfolds, at least partially. Recent work used MD simulations to study the conformational dynamics of insulin in solution and identified several major “elements of disorder” to describe the partially unfolded intermediate structures they observed (Busto-Moner et al, 2021). We docked these structures of insulin into the closed catalytic chamber of IDE and found that the chamber was able to easily accommodate the positioning of either chain near the exosite and catalytic cleft in cis and trans orientations (Figure 2 – figure supplement 3).

Heterogeneity analysis reveals 2 dominant components of structural variability with the IDE cryo-EM particle population

Numerous structures of IDE exist with subunits adopting either an open or closed conformation, yet we lack an understanding of how IDE transitions from an open to a closed state. The simplest explanation of this transition, based on structural data, would be a direct, rigid body translation of IDE-N relative to IDE-C, and this is the predominant model within the field (Shen et al., 2006; Zhang et al., 2018). To better understand what this transition would look like, we measured and plotted the changes in the D1-D4 COM distance versus the changes in the D1-D2-D3-D4 COM dihedral angle for all available cryo-EM structures of IDE (Figure 3 – figure supplement 1). We found that in the absence of substrate, the transition pathway produced a linear relationship between states. However, when our structures generated in the presence of insulin were analyzed, the linear relationship no longer held (Figure 3 – figure supplement 1), suggesting that the simple linear transition typically presumed for the open-close transition between distinct IDE states does not accurately depict the complexity of IDE dynamics.

Recently, several approaches have been developed to understand the conformational heterogeneity present within cryo-EM data. We employed multibody analysis in RELION and 3D variability analysis (3DVA) in cryoSPARC to investigate the conformational heterogeneity within our particle populations (Nakane et al, 2018; Punjani & Fleet, 2021). Of these, the range of motion predicted by multibody analysis most closely matched the reported structures while 3DVA exhibited consistent results across a smaller magnitude of transition (Figure 3 – figure supplements 1-3). Therefore, we focused on multibody analysis below. Multibody analysis models the structural heterogeneity in our data as the result of motions of independent, user-defined rigid bodies. As discussed above, the primary source of structural variation in our structures is the positioning of IDE-N relative to IDE-C. We see little structural changes within the domains, suggesting that IDE domains behave as three rigid bodies, IDE-N in two different subunits within IDE dimer (IDE-N A, IDE-N B) and the dimeric IDE-C (IDE C/C). This is consistent with the structural analysis of M16 family of metalloproteases (Liang et al., 2022). This assertion is further supported by our all-atom molecular dynamics (MD) simulations. While we observed small yet noticeable conformational changes within IDE-N A, IDE-N B, and IDE-C/C, the overall RMSD of entire IDE dimer is significantly higher than those three individual units (Fig. 3A). Multibody analysis defines the principal components of variance along discrete eigenvectors, which we interpret as proxies for the dominant components of molecular motion. We quantified the change in D1-D4 COM distance and D1-D2-D3-D4 COM dihedral of representative structures along the heterogeneity gradients of the top 9 eigenvectors for each of our IDE structures, representing ∼70-90% of the total structural heterogeneity per structure. The results indicated that the particles comprising our cryo-EM structures sampled a significantly greater conformational space than would be expected from analysis of the ensemble structures alone (Figure 3 - figure supplement 2). Most intriguingly, the conformational changes displayed an unexpectedly high degree of change in the D1-D2-D3-D4 COM dihedral angle, indicating a significant rotation of IDE-N relative to IDE-C (Fig. 3B,C, Figure 3 – figure supplement 2).

Conformational dynamics of IDE implied by structural heterogeneity.

(A) All-atom MD simulations analysis. The primary source of structural variance (RMSD) results from the IDE-N moving against IDE-C as a rigid body. Rigid bodies were defined as colored for multibody refinement in RELION. (B-C) Multibody analysis. The range of conformational variance described by the top principal component vectors displays an unexpectedly high degree of rotational motion, as measured by the change in D1-D2-D3-D4 dihedral angle across the gradient of structural heterogeneity for each vector, in both the absence (B) and presence (C) of insulin compared to the expected open-close transition pathway predicted from a linear interpolation of the experimentally determined structures of IDE (dashed line, Figure 3 – figure supplement 1). Two dominant components of structural variance are revealed from multibody analysis: (D) where IDE-N swings relative to IDE-N about the inter-domain linker, and (E) where IDE-N rotates against IDE-C. Starting (orange) and ending (red) states of IDE-N shown with pathway depicted by arrows. IDE-C shown as gray surface.

Integration of the multibody results across all states of IDE suggest two dominant components of structural variance. First is a translation-dominant conformational change wherein IDE-N swings toward or away from IDE-C as if mediated by a hinge formed by the interdomain linker region of IDE (Fig. 3D, Figure 3 - video 1). This conformational change closely resembles our understanding of the IDE open-closed transition inferred from analysis of the ensemble structures. The second component is a rotation-dominant conformational change wherein IDE-N rotates orthogonal to the plane of the dimer as if it were being screwed into or ground against IDE-C (Fig. 3E, Figure 3 – video 2). Importantly, these dominant components of structural variance were also observed when the particle populations were analyzed with 3DVA in cryoSPARC, albeit over a smaller magnitude of conformational change (Figure 3 – figure supplement 3). These dominant components also correlate well with the lowest frequency modes revealed by normal mode analysis (Figure 3 – videos 3, 4). Interestingly, while the presence of insulin was found to substantially influence the consensus reconstructions of IDE, we found that the presence or absence of insulin had no significant effect on the principal components of structural variance for IDE, consistent with the current understanding that enzyme conformational changes are “hardwired” into the structure (Bahar et al, 2007; Bahar & Rader, 2005).

R668 plays a key role in the IDE open/close transition in all-atom MD simulations

Analysis of the particle heterogeneity comprising cryo-EM structures provides excellent information about the conformational space sampled by the protein of interest. However, such analysis lacks a temporal component, and thus only implies motions. To overcome this barrier and examine how the conformation of IDE changes over time, we performed six replicate all-atom MD simulations with explicit solvent. Simulations were initialized from the O/pO state (PDB: 7RZG), with the missing loops modeled in to generate continuous peptide chains and run for one microsecond each. By starting with the O/pO structure, we investigated the dynamics of both the open and closed subunits simultaneously. We observed that the conformational space sampled by our simulations correlated well with the conformational space sampled by our cryo-EM particle population, as revealed by multibody analysis (Figure 4 – figure supplement 1). In five of our six simulations, the open subunit closed quickly, typically in <200 ns (Fig. 4A). Analysis of the open-close transition revealed two key findings. First, the open subunits did not close to a singular consensus structure. We define closing as a subunit reaching a similar D1-D4 COM distance as the pO or pC states, but we observed distinct differences among the subunits indicative of IDE-N rotation relative to IDE-C. These differences are in line with our second key finding, that the conformational changes associated with the open-close transition did not follow the same general pathway among the simulations, as we observed a high degree of variability upon closing (Fig. 4B).

All-atom MD reveals a molecular basis for IDE conformational dynamics.

(A) Measurements of the O subunit D1-D4 distance over the course of six separate microsecond long all-atom MD simulations of WT IDE. Of which, the open subunit closed in 5 of the 6 simulations. (B) Plot of the O subunit D1-D4 distance vs the D1-D2-D3-D4 COM dihedral angle over the course of the simulation of WT IDE. The open subunits displayed a variety of closing pathways and did not close to a consensus structure. Starting structure shown as black dot, pO structure shown as white dot. (C) R668 acts as a guidepost residue, rapidly interacting with D309 or E381. Formation of this interaction is associated with rapid closing, as measured by a decrease in D1-D4 distance (D). (E) Hydrogen-deuterium exchange mass spectrometry highlights the importance of R668 in mediating the open-close transition. In the presence of insulin (panel 1, red), Aβ (panel 2, red), and BDM-44768 (panel 3, red), all of which promote IDE closing, the peptide containing R668 shows reduced deuterium exchange relative to apo-IDE (black), yet in the presence of 6bk (panel 4, red), which does not promote closing, there is no difference in the exchange rates for the R668 containing peptide relative to apo-IDE (black). Helix containing R668 colored by red (increase) – white (no change) – blue (decrease) gradient depicting the degree of deuterium exchange relative to apo-IDE. (F) Measurements of the O subunit D1-D4 distance over the course of six separate microsecond long all-atom MD simulations of IDE R668A. (G) Plot of the O subunit D1-D4 distance vs the D1-D2-D3-D4 COM dihedral angle over the course of the simulation of IDE R668A. The six separate microsecond long simulations indicate that an R668A mutation significantly alters the closing dynamics of IDE (F) and increases the rotational motion (G) relative to WT (panels A and B respectively). Starting structure shown as black dot, pO structure shown as white dot.

Noticeably, R668 consistently stood out as one of the first IDE-C residues to form new interactions with IDE-N, either D309 or E381 despite the path variation among our MD simulations (Fig. 4C). In most simulations, formation of this interaction preceded a large decrease in the D1-D4 COM distance (Fig. 4D, Figure 4 – figure supplement 2). Interestingly, these R668 interactions are not present in the ensemble cryo-EM or crystal structures of IDE. We hypothesize that the biochemical properties and spatial localization of R668 enable it to essentially reach out and grab onto the N-domain, initiating the formation of a complex hydrogen bonding network that drives the open-close transition towards completion. Previously published hydrogen-deuterium exchange mass spectrometry (HDX-MS) data indicates that R668 is stabilized under conditions that promote IDE closing, suggesting that R668 may function as a molecular latch (Fig. 4E) (Zhang et al., 2018).

To further probe the importance of R668, we ran 6 additional simulations with IDE carrying a R668 to alanine (R668A) point mutation under identical conditions as described previously. These simulations displayed significantly altered dynamics when compared to the wild-type simulations. Notably, the R668A mutant was found to close more slowly, and many simulations did not reach the D1-D4 COM distance typified by a closed state structure, instead stabilizing with a significantly larger D1-D4 COM distance than we had previously observed in our wild type (WT) simulations or our experimental structures (Fig. 4F). We also observed that the R668A mutant preferentially sampled a subset of conformational space rarely explored in our WT simulations and displayed significantly greater variation in the D1-D2-D3-D4 dihedral angle (Fig. 4G). As a result of this increased rotational exploration, several of our R668A simulations closed to a conformation where IDE-N is offset and rotated relative to IDE-C when compared to the experimentally determined structures (Figure 4 – figure supplement 3). This altered conformation results in a structure where IDE-N and IDE-C are fully engaged, but the offset produces a deceivingly high D1-D4 COM distance. With this in mind, we sought to better understand the conformational dynamics of IDE in the closed state.

R668A modulates IDE conformational dynamics in vitro

The R668A mutation was found to significantly alter the conformational dynamics of the IDE open-close transition in our simulations (Fig. 4F, G), so we expressed and purified the R668A mutant to determine if the changes predicted from our MD simulations resulted in altered enzymatic activity relative to WT. Previous work has established that IDE adopts a dominant O/pO state in solution (Zhang et al., 2018). Consistent with our MD simulations, size-exclusion chromatography revealed that the R668A construct eluted slightly earlier than WT IDE, suggesting that the R668A mutation induces a greater proportion of molecules to adopt a more open conformation, possibly shifting to a dominant O/O state (Fig. 5A). Next, we compared the enzymatic activity of the R668A and WT constructs using the fluorogenic substrate (7-methoxycourmarin-4-yl)acetyl-RPPGFSAFK(2,4-dinitrophenyl)-OH (substrate V), a bradykinin mimetic which has previously been used to characterize the enzymatic activity of IDE and related enzymes (King et al, 2014; McCord et al., 2013). We found that WT IDE has activity consistent with previously published data (McCord et al., 2013) and the R668A mutation produced a ∼5-fold decrease in activity (Fig. 5B). ATP is known to promote IDE degradation of small substrates, i.e. bradykinin, but not larger substrates, i.e. insulin (33). Consistent with this observation, the addition of ATP increased the activity of both WT and R668A constructs (Fig. 5B). We also tested if the R668A mutation would alter the degradation of larger, well-folded substrates, using insulin in a direct competition assay. We found that, after accounting for the previously observed 5x decrease in activity, insulin yielded an apparent Ki of ∼8 nM for WT IDE, but this value was reduced to ∼52 nM for the R668A construct (Fig. 5C). When Aβ(1-40), an unstructured substrate, was tested in the same competition assay, we observed an apparent Ki of ∼688 nM for WT IDE, yet the R668A mutant did not exhibit a dose-dependent response to the presence of Aβ and the data could not be fit to any kinetic equation (Fig. 5D). We then performed size exclusion chromatography-coupled small-angle X-ray scattering (SEC-SAXS) for both constructs to ascertain if the observed differences in enzymatic activity could be explained by altered biophysical properties. The WT scattering profile produced an Rg of 49.4 +/-0.9 Å, consistent with previous SEC-SAXS results (Fig. 5E, F) (Zhang et al., 2018). The R668A scattering profile produced an Rg of 54.7 +/- 0.4 Å, consistent with our SEC data that R668A mutant has slightly larger hydrodynamic radius (Fig. 5A). Together, our data indicates that R668A mutation can profoundly affect the conformational dynamics governing the open-closed transition, and that altered conformational dynamics can alter enzymatic activity in a substrate-dependent manner.

R668A alters IDE activity in vitro.

(A) Elution profile of WT IDE (blue) compared to the R668A mutant (orange) from a S200 SEC column. (B) Degradation of the fluorescent substrate MCA-RPPGFSAFK(Dnp) by WT IDE and the R668A construct in the presence and absence of ATP. Data represents the average initial velocities of three replicates performed at a protein concentration of 3.125 nM. Error bars (gray) represent the standard error. (C) Inhibition of MCA-RPPGFSAFK(Dnp) degradation by WT IDE (circles, solid fit lines) and IDE R668A (squares, dashed fit lines) in the presence of varying amounts of insulin. Data was fit to the Michaelis-Menten (black) and Hill equations (red). Relevant parameters, Michaelis-Menten: WT: χ2=0.001, Vmax=0.951, Ki=8.3 nM; R668A: χ2=0.005, Vmax=0.892, Ki=52 nM; Hill: WT: χ2=0.009, n=0.55, Ki=51 nM; R668A: χ2=0.055, n=0.61, Ki=198 nM. Error bars represent standard error, data points represent the average of three technical replicates. (D) Inhibition of MCA-RPPGFSAFK(Dnp) degradation by WT IDE (circles, solid fit lines) and IDE R668A (squares, dashed fit lines) in the presence of varying amounts of Aβ1-40. WT data was fit to the Michaelis-Menten (black) and Hill equations (red), R668A data could not be fit to either equation, indicating that the mutation confers substrate-specific altered enzyme kinetics. Relevant WT parameters, Michaelis-Menten: χ2=0.066, Vmax=0.823, Ki=353 nM; Hill: χ2=0.08, n=0.72, Ki=688 nM. Error bars represent standard error, data points represent the average of three technical replicates (E) SEC-SAXS profile of WT (black) and R668A (red) constructs with Rg values calculated by both the Guinier and Porod methods along with Dmax derived from the P(r) function (F).

Charge-swapping at the IDE-N/C interface mediates conformational dynamics in the closed state

The WT pO subunit remained closed throughout all simulations yet was found to initially relax to a slightly more open state within the first 200 ns, roughly the same time frame it took the WT O subunit to close (Fig. 6A) and, despite not opening, demonstrated a range of D1-D2-D3-D4 dihedral variation similar to that of the O subunits (Fig. 6B). Interestingly, the conformational variation within our simulations increased greatly after this relaxation point was reached (Fig. 6A). This relaxed conformational state consistently displayed an altered IDE-N/C interface compared to the experimentally determined structures that was replicated by the open WT subunits upon closing, leading to a nearly symmetric pO/pO state (Figure 6 – figure supplement 1). Biochemically, this relaxed IDE-N/C interface makes more sense than the interface observed experimentally. It has previously been demonstrated that the experimental constraints of crystallographic structure determination force IDE to adopt a closed conformation and determination of the open state can only be accomplished once those restraints are removed, i.e. cryo-EM (Zhang et al., 2018). These results suggest that both ensemble methods of structure determination may impact the conformation of IDE, however slightly. The IDE-N/C interface is littered with charged residues, yet few interactions are observed in the ensemble structures (Fig. 6C), whereas our MD simulations reveal a complex hydrogen bonding network (Fig. 6D). We observed an extensive network at the D2-D3 interface, with sporadic patches of interactions between D1 and D4 (Figure 6 – figure supplement 2). Within the D2-D3 network R668 again stands out as a notably key residue. In addition to forming hydrogen bonds with D309 and E381, we observed R668 form τ-τ interactions with R311 (Fig. 6E). Arginine-mediated τ-interactions have been well characterized in other systems for their ability to stabilize interaction interfaces and facilitate conformational changes (Armstrong et al, 2016; Vernon et al, 2018).

Structural basis of closed state conformational dynamics.

(A) Measurements of the pO subunit D1-D4 distance over the course of six separate microsecond long all-atom MD simulations of WT IDE. (B) Plot of the pO D1-D4 distance vs the D1-D2-D3-D4 COM dihedral angle over the course of the simulation of WT IDE. (C) IDE-N/C interface previously solved crystal structures (PDB:2G47 shown) shows side chains are ill-positioned for interaction. (D) IDE-N/C interface formed upon open subunit closing in our MD simulations reveals a complex hydrogen bonding network. (E) Heat map showing conformational geometries that were preferentially sampled in our MD simulations by the open subunits upon closing. Insets highlight how the IDE-N/C interface changes to permit interdomain motion. (F) Plot of the O subunit D1-D2-D3-D4 dihedral angle during a subset of a single WT IDE MD simulation after the open-close transition has been completed. Charge-swapping between residues at the IDE-N/C interface is associated with changes in the D1-D2-D3-D4 dihedral. (G) For most of the simulation, D309 interacts with K483 (black), however, this interaction is broken for ∼100 ns, during which D309 instead interacts with R311 (blue) and R668 (orange). (H) For most of the simulation, D426 interacts with K571 (black), yet this interaction is periodically broken, and D426 instead interacts with K425 (green) and K899 (magenta). When these events of charge-swapping coincide with D309 charge-swapping (G), they are associated with a large change in the D1-D2-D3-D4 dihedral angle (F). When they occur alone, the effect on D1-D2-D3-D4 dihedral is smaller.

The IDE-N/C interactions we observe in our simulations are not restricted to discrete cognate pairs; rather, we observed that residues within the network periodically swap among several interaction partners. This “charge-swapping” phenomenon allows IDE-N to rotate or slide against IDE-C to adopt multiple conformations and maintain favorable interacting contacts (Fig. 6F). While multiple, concurrent, instances of charge-swapping were associated with rapid changes in the D1-D2-D3-D4 dihedral angle, singular events of charge-swapping did not appear to be significant enough to stimulate large-scale conformational change (Fig. 6G, H). Instead, substantial conformational changes appear to be caused by multiple charge-swapping events occurring in conjunction. This may explain why we observed no noticeable change in the conformational dynamics of the pO state subunits when comparing the WT and R668A constructs. Consurf analysis reveals that most of the residues forming these interaction networks are highly conserved among IDE homologs (Figure 6 – figure supplement 3) (Ben Chorin et al, 2020). This indicates that the revealed rotational motion is likely evolutionarily conserved and offers a potential mechanism by which IDE unfolds and repositions bound peptide substrates to degrade amyloid peptides.

Time-resolved cryo-EM suggests insulin-induced allostery in IDE

To begin understanding of how the interaction between IDE and insulin leads to insulin unfolding and degradation, we first performed all-atom MD simulation of insulin-bound IDE in the closed state. Unfortunately, we did not observe a noticeable structural change of insulin within our microsecond simulations. Previously, we used time-resolved SAXS analysis to determine the time constant for the insulin induced open to closed transition, tau (1, a.k.a., monoexponentially decay) to be ∼100 milliseconds (Zhang et al., 2018). However, it has previously been determined that IDE cleaves about two insulin molecules every second (Manolopoulou et al., 2009). We employed spotiton to rapidly mix and freeze grids where IDE has been exposed to insulin for 123 milliseconds prior to freezing (Dandey et al, 2020). We hoped that such an experimental setup would permit the identification of novel insulin-bound states of IDE and offer insight into the initial binding and unfolding of insulin. Previous work has established that, in the absence of substrate, IDE adopts a dominant O/pO state in solution and the presence of a molar excess of substrate shifts the particle population to a dominant pC/pC state (Zhang et al., 2018). Surprisingly, after processing the dataset in Relion, we observed a single conformational state within our rapidly-mixed dataset: the O/O state (Fig. 7A, Figure 7 - figure supplement 1, supplementary file 2). We were unable to resolve the structure beyond moderate resolution (∼5.1 Å) and did not observe unambiguous density indicative of insulin binding within the catalytic chamber, which is unsurprising given our previous observations of promiscuous binding (Fig. 2C). Despite the relatively high particle number in our final reconstruction, we were unable to sub-classify distinctly different conformations of IDE within the particle population as initial classes converged back to the dominant O/O state upon subsequent refinement. This leads us to believe that the obtained O/O state represents the average state of a particle population that exhibits a high degree of continuous structural heterogeneity. The O/O state observed in our rapidly-mixed dataset is distinct from the previously observed O/O states with D1-D4 center of mass distances of 55.1 Å and 46.8 Å matching the largest and smallest, respectively, distances previously observed for individual O state subunits (Fig. 7B). The discrepancy between the “openness” of the two subunits nearly doubles that observed in previous structures, making this O/O state the most asymmetric O/O state yet observed (Fig. 7C). The observation of a dominant O/O state in the presence of insulin under short timescales suggests that the binding of insulin to the O subunit of the apo-dominant O/pO state can induce the pO state to open through an unknown allosteric mechanism to reach the observed dominant O/O state (Fig. 7D). Positive allostery in IDE has previously been identified in biochemical assays, yet we lack a detailed understanding of the underlying mechanism (McCord et al., 2013; Song et al, 2011).

Time-resolved cryo-EM of IDE+insulin reveals a new O/O state.

(A) 5.1 Å reconstruction of IDE+Fab rapidly mixed with a 5x molar excess of insulin and vitrified with a mix-to-freeze time of 123 milliseconds. (B) Measurements (as in Fig. 2) of the distance between the D1-D4 centers-of-mass and dihedral angle formed by the D1-D2-D3-D4 centers-of-mass as indicators of the “openness” of the structure depicted in (A). (C) Alignment of the O/O state from our rapid-mixing dataset (domains colored as in Fig. 1) compared to the previously solved O/O states (gray, as labeled). (D) Overview of the conformational states adopted by IDE in all 4 available cryo-EM datasets with corresponding percentage of particles mapping to each state (Zhang et al., 2018)(this work).

Coarse-grained MD simulations reveal intermolecular cross-β sheet formation between insulin and IDE in all four IDE domains

As cryo-EM proved unsuitable to explore the mechanism of insulin unfolding by IDE, we turned to MD simulations. As stated above, there is a notable discrepancy in the time frame between the open/close transition and insulin cleavage rate. This suggests that unfolding requires multiple cycles of the open-closed transition and, thus it is not currently realistic to use all-atom MD simulations to probe the unfolding of insulin by IDE. We then used a coarse-grained MD package, Upside, to probe the interaction between IDE and insulin. Upside performs the dynamics simulation for just six atoms per residue with the sidechains being represented by directional-dependent beads whose packing probabilities have the lowest free energy while still including key structure details to compute realistic forces (Faruk et al, 2022; Faruk et al, 2023; Jumper et al, 2018a, b). This allows rapid equilibration and sampling of the energy landscape when compared with the conventional all-atom MD simulation, achieving ∼104 acceleration for the calculation time. Consistent with all-atom MD simulations, Upside simulations revealed the wide range of motions, including grid and hinge motions found in our all-atom MD simulation (Figure 8 – figure supplement 1). We observed the partial unfolding of insulin in our Upside simulations yet, importantly, the unfolding did not disrupt the inter-or intra-molecular disulfide bonds (Figure 8 – video 1).

Upside analysis of IDE-insulin interactions.

(A) Transient cross-β-strand interactions between insulin and IDE. Secondary structure (gray: coil, green: β-strand, yellow: α-helix) for each insulin residue (Y-axis) over the course (X-axis) of a representative Upside simulation. Points indicate predicted hydrogen bond interactions between IDE and insulin, colored by which domain of IDE that insulin residue is interacting with (D1: red, D2: orange, D3: blue, D4: cyan). Stars below the X-axis indicate frames from which structures were extracted for panel B. (B) Structure of IDE/insulin complex in Upside simulations at frames indicated in (A). IDE domains colored as in (A), insulin shown in magenta (chain A) and yellow (chain B). Inset highlights the intermolecular cross-β sheets formed between IDE and insulin within each of the four homologous IDE domains. (C) Individual domains of IDE, colored as in (A), extracted from (B) and oriented to demonstrate the structural conservation of intermolecular cross-β-sheet formation with bound insulin (colored as in B).

As insulin partially unfolds, it readily adopts β-strand structures—a key characteristic of amyloidogenic peptides—which can transiently form cross-β-sheets with the central β-sheet present in all four domains of IDE (Figure 8A-B). Specifically, we observed that cross-β-sheet interactions occur at distinct regions of insulin: the N-terminus of insulin chain A interacts with the IDE D2 domain near the substrate N-terminal anchoring site (also known as the exosite); the middle portion of insulin chain A interacts with the IDE D1 domain; and the C-terminal regions of insulin chain B interact with the D3 and D4 domains of IDE. While the interactions of insulin with IDE-C domain are novel, those with IDE D1/D2 domains are structurally similar to those previously observed (Guo et al, 2010; Manolopoulou et al., 2009; Ren et al, 2010; Shen et al., 2006). The labile and transient β-strand-rich interactions have been shown to mediate diverse protein-protein interactions, e.g., the assembly of intermediate filaments (Eibauer et al, 2024; Zhou et al, 2021).

This interaction provides molecular insight into how IDE recognizes and binds amyloidogenic peptides with diverse sequences (Figure 8C). IDE is capable of forming cross-β interactions with many regions of insulin, primarily mediated by backbone-to-backbone contacts that do not rely on specific amino acid sequences. The structural arrangement of the four conserved domains of IDE enables it to interact with multiple regions of its substrates, such as insulin, which are predisposed to forming β-strands (Figure 8C). We term this type of interaction “β-grabbing,” and propose that it is a key mechanism by which IDE unfolds insulin. This interaction effectively tethers the substrate to both the IDE-N and IDE-C domains, facilitating the interdomain motions of IDE that promote substrate unfolding.

Discussion

Our results allow us to put forth a refined model describing how IDE recognizes amyloid peptides that are diverse in size and shape (Fig. 9). The catalytic cycle of IDE starts with at least one subunit of dimeric IDE adopting the open state, exposing the interior of the catalytic chamber to potential substrates (Zhang et al., 2018). The unconstrained motion of IDE-N likely facilitates substrate interaction in a variety of initial orientations. IDE-N and IDE-C have negatively and positively charged surfaces, respectively, and are able to attract peptide substrates with complementary charge profiles (Shen et al., 2006; Zhang et al., 2018). Such peptides have a high dipole moment and are often prone to aggregation. However, in the open state, the door subdomain, which contains the key catalytic residues, is highly flexible, rendering the open state catalytically incompetent (McCord et al., 2013; Zhang et al., 2018). Proper positioning and unfolding of substrate is necessary to stabilize the cleft prior to catalysis. Substrate binding has been suggested to enhance IDE closing through charge complementarity (Shen et al., 2006). It is likely that the rotation of IDE-N permits a range of closing pathways to accommodate non-optimal substrate binding geometry (Fig. 4). If the closing geometry permits R668 to interact with E381 or D309, the closing reaction continues to completion. If the closing geometry does not permit the R668-mediated interactions, this could be a signal of improper substrate binding, allowing IDE-N to open and either release improper substrate or reposition for another closing attempt. Once closed, IDE-N is capable of rotating against IDE-C mediated by charge-swapping events at the IDE-N/C interface (Fig. 6). The combined motions associated with the open to closed transition and rotation between IDE-N and IDE-C can lead to the distortion of the secondary structure of bound substrate. Furthermore, the transient interaction of insulin with the catalytic chamber of IDE such as β-grabbing via the β−strand pairing should further stress the integrity of bound substrate (Fig. 8). Together, these forces should facilitate unfolding and proper positioning of substrate for cleavage. Our data indicates that IDE is able to accommodate any possible orientation of insulin tethering within the catalytic site (Fig. 2C), suggesting that IDE is able to cut both chains of insulin in rapid succession without requiring significant repositioning (Manolopoulou et al., 2009). IDE then opens to release cleavage products and the cycle begins anew.

Model for the catalytic cycle of IDE.

The details are described in discussion. For simplicity, only the scenario involving the addition of a second insulin molecule is illustrated, although this is not required for the catalytic cycle. The mechanism by which IDE transitions between closed and open states remains unknown and is therefore indicated with question marks.

IDE utilizes both unfoldase and protease activities to degrade clinically-relevant peptides, including three glucose-regulating hormones with contrary effects – insulin, amylin, and glucagon – and Aβ, the accumulation of which is associated with the progression of Alzheimer’s disease. This has made IDE an attractive target for therapeutic intervention (Maianti et al., 2019; Tang, 2016). One major challenge is how to better control selectivity when the substrates IDE degrades are highly diverse in sequence and structure. It remains unknown how IDE unfolds substrates prior to cleavage. Our studies characterize the conformational dynamics of IDE and offer insight into the catalytic mechanisms, revealing that IDE-N rotates against IDE-C. We envision that this motion works in concert with the open-close transition to reposition and unfold substrates prior to cleavage. This rotation appears to be largely unconstrained in the open state and may play a role in promoting substrate capture. In the closed state, we found the rotational motion to be mediated by charge-swapping events at the IDE-N/C interface. These events are supplemented by interactions between hydrophobic patches that provide a non-specific interaction interface and allow the domains to slide against one another. We found that IDE-N/C interactions were mediated by the key residue R668, and mutation of this residue altered the enzymatic activity and conformation of IDE, likely by impairing the open-close transition. Our data suggests that, while the R668A mutation globally impaired IDE activity, R668A exhibits significantly different kinetic profiles towards well-folded substrates (insulin) compared to unstructured substrates (Aβ). This highlights the complexity of the IDE catalytic cycle as we envision that IDE conformational dynamics can regulate solvent access to the catalytic chamber and substrate binding, unfolding, and cleavage are all intimately linked.

Our work adds to the expanding body of literature that integrates computational and experimental approaches (Arenz et al, 2016; Hirschi et al, 2021; Jiang & Doudna, 2017; Palermo et al, 2016; Roh et al, 2020; Saha et al, 2022). By employing multibody analysis, we identified the dominant sources of structural heterogeneity within our cryo-EM particle population, uncovering that the conformational dynamics of IDE involve a substantial rotational component that is not evident from ensemble structure analysis alone. We showed that all-atom MD simulations can address limitations of cryo-EM single particle analysis when investigating the role of residue R668. Because R668 forms a network of polar and electrostatic interactions rather than exclusive partnerships, dynamic hydrogen bonding networks at the IDE-N/C interface—particularly those involving R668—contribute to conformational heterogeneity, which further restricts the achievable resolution in cryo-EM analysis. Furthermore, vitrification by freezing proteins in liquid ethane can significantly alter the entropic contribution of R668 through conformation-dependent promiscuous contacts. By combining cryo-EM and MD simulations, we gained insight into how flexible interactions between IDE-N and IDE-C facilitate domain movements and highlighted the critical role of R668 in mediating these interactions. We also used MD simulations to demonstrate altered open-closed dynamics in the R668A mutant compared to the wild-type enzyme and validated these findings with experimental enzymatic kinetic assays and SEC-SAXS experiments.

At its core, cryo-EM is an ensemble method of structure determination, averaging tens of thousands of particles together to generate consensus structures. While methodologies have been well-established to separate out discrete conformational classes, proteins that exhibit constant gradients of structural heterogeneity remain problematic, although significant attention has been devoted to the issue in recent years, with varying levels of success (Nakane et al., 2018; Punjani & Fleet, 2021; Zhong et al, 2021). Currently, the best approaches use various dimensionality reduction techniques to analyze the principal components of particle heterogeneity. While a substantial innovation, it remains to be seen how the principal components of structural heterogeneity, derived from a series of static snapshots of a particle at the time of freezing, correlate with the real molecular motions of a protein in an aqueous environment. For many systems, especially enzymes of clinical significance, proper characterization of the conformational dynamics and transitional states of the protein are necessary for a full understanding of protein function and lays the foundation for future therapeutic development. MD simulations can provide this information, yet such studies are often met with skepticism by experimental researchers. By integrating MD simulations with experimental techniques, the weaknesses of one technique can be compensated for by the strengths of other techniques to provide comprehensive information about the system of interest. While this integrative approach has proven beneficial across a wide range of systems, including proteorhodopsin, the ribosome, and rotary ATPases, perhaps the best example of this approach is the CRISPR-Cas9 system, where an exhaustive number of studies have combined cryo-EM, crystallography, and traditional biophysical techniques with traditional and accelerated MD methods to drive the field forward at an astounding pace (Arenz et al., 2016; Hirschi et al., 2021; Jiang & Doudna, 2017; Palermo et al., 2016; Roh et al., 2020; Saha et al., 2022).

Our integrative structural analyses provide the structural basis for IDE conformational dynamics and their roles in IDE catalytic cycle and the framework for future studies. We speculate that multi-cycles of open-close transition and rotation driven by IDE conformational dynamics will be required for its unfoldase activity. However, the kinetic timescale of this reaction is fast, estimated to be ∼500 milliseconds rendering investigation challenging experimentally while the timescale is long for all-atom MD simulation (Manolopoulou et al., 2009). Our cryo-EM analyses also shed light on IDE allostery (McCord et al., 2013; Song et al., 2011; Zhang et al., 2018) (Fig. 9). The IDE dimer primarily adopts the O/pO conformational state (Zhang et al., 2018). Through time-resolved cryo-EM analysis, our data suggest that, upon substrate binding to the O state, the substrate-inaccessible pO state undergoes allosteric opening, allowing this subunit to capture its substrate. Additionally, the prolonged incubation of IDE with sub-stoichiometric amounts of insulin results in the predominance of the pC/pC state, suggestive of coordination in substrate capture and open-closed transition. Collectively, these positive allosteric transitions would enhance the capacity of IDE to capture and degrade its substrates more efficiently, though the underlying mechanisms remain unclear. Future integrative structural studies are required to further elucidate and control the IDE catalytic cycle, enabling substrate-specific modulation of IDE activity and advancing its therapeutic potential.

Materials and Methods

Expression and purification of human IDE

Cysteine-free IDE (referred to as IDE in this study) was expressed in E. coli BL21 (DE3) cells (at 25 °C for 20 h, 0.5mM IPTG induction using T7 medium). The R668A mutation was introduced into this construct by site-directed mutagenesis with the following primer pair: CCGGAAATTGTTAAGAGATGCCATATATGCTTCTTTG (forward) CAAAGAAGCATATATGGCATCTCTTAACAACCGGGC (reverse) and verified by sequencing. Recombinant IDE proteins were purified by Ni-NTA, source-Q, and Superdex 200 columns as previously described (Zhang et al., 2018). Protein was aliquoted and stored at −80° C.

Cryo-EM data collection and analysis

Thawed IDE aliquots were further purified by Superdex 200 chromatography using buffer containing 20 mM HEPES, pH 7.2, 150 mM NaCl, 10mM EDTA and then mixed with FabH11-E at an equal molar ratio. FabH11-E-IDE complex was purified by Superdex 200 chromatography in the presence of insulin at molar ratio IDE:insulin=2:1. Insulin was purchased from SIGMA (91077C). All grids, 300 mesh carbon or gold holey nanowire grids were plasma cleaned with O2 and H2 for 10 secs using a Solarus plasma cleaner (Gatan) and plunged at 133 milli-second using Chameleon, Spotiton-based technology. All images were acquired using a Titan Krios microscope (FEI) operated at 300KeV with a Gatan K3 direct electron detector (Gatan) in counting mode. Images were automatically acquired using Leginon (Suloway et al, 2005). Images were processed using software integrated into RELION3 (Zivanov et al, 2018). Frames were aligned using MotionCor2 (Zheng et al, 2017) software with dose weighting, CTF was estimated using Gctf (Zhang, 2016), particles were picked and extracted automatically using RELION. Particle stacks were processed through several rounds of 2D and 3D classification. Selected classes were then processed for high-resolution 3D refinement. The flowchart and detailed data processing is summarized in Figure 2 – figure supplement 1. Finally, the overall map was improved by particle polishing in RELION and sharpening. The final resolution was estimated using Fourier Shell Correlation (FSC=0.143, gold-standard). The density fitting and structure refinement was done using UCSF CHIMERA (Pettersen et al, 2004), COOT (Emsley & Cowtan, 2004), REFMAC5 (Murshudov et al, 2011) and PHENIX (Adams et al, 2010). Time-resolved cryo-EM grids by rapidly mixing IDE+Fab with insulin were prepared using the Spotiton device as previously described (Budell et al, 2021; Dandey et al., 2020). In brief, IDE+fab and insulin were thawed on ice and diluted in 20 mM HEPES, pH 7.2, 150 mM NaCl, 10 mM EDTA to 25 µM and 127.5 µM, respectively. Self-wicking, nanowire grids were plasma treated (Gatan Solarus) at 5 W in hydrogen and oxygen for 1-2 minutes (Wei et al, 2018). Each sample was aspirated into separate piezodispensers and applied sequentially in 70-100% relative humidity at 22-25C onto plunging grids just before cryogen entry. The maximum constant velocity of the plunging robot was adjusted to obtain the target sample mixing times of 123 milliseconds. The data was collected and processed as above, in Relion5, with Blush regularization being employed during 3D refinement.

Multibody analysis

We defined three rigid bodies as shown in figure 2A for the analysis. The size of the user-defined bodies has been theorized to play a significant role in the success of multibody refinement, most likely due to the fact that larger bodies will have a stronger signal-to-noise ratio (Nakane et al., 2018). With a Fab bound to each of IDE-N within IDE dimer, we assessed the impact of body size on multibody refinement by examining how density corresponding to 2 Fv regions, 1 Fv region, or no Fv regions would affect the multibody refinement. Fab has a variable region (Fv), and a constant region (Fc) and we chose to subtract and mask out the density corresponding to the Fab Fc, as we speculated that any motion between the Fab Fv and Fc domains would bias the results of the multibody refinement. The quality of the multibody results was assessed based on the subsequent improvement in the Coulomb potential map quality and calculated resolution. In the absence of any Fv density, the maps resulting from multibody refinement had the best calculated resolution, yet the map quality was quite poor overall; much of the density appears globular and featureless, particularly in the IDE-N regions (Figure 2 – figure supplement 4). Conversely, when Fv density was present on both IDE-N bodies, the resolution of the multibody output maps (4.5 Å) was worse than the resolution of the input map (4.3 Å). Thus, we found that the greatest improvement occurred when the Fv density was present on only the exterior body of the pO subunit. In this case, multibody refinement improved both the calculated resolution and density quality.

Following multibody refinement, the structural heterogeneity for each body was analyzed along six principal component vectors, resulting in 18 vectors describing the primary components of structural variation within the IDE dimer. We analyzed the variation along these vectors as a proxy for protein motion. The results of the multibody analysis indicate the degree of structural variation that is explained by motion along each of the primary component vectors. For all IDE states, we observed no single component vector could explain the majority of structural variance (Supplementary files 3-8). Given the underlying assumptions of multibody refinement, there is a question of whether any observed motions are biologically relevant or simply the result of improper particle alignment. We reasoned that the greater the degree of variation explained by motion along a specified component vector, the greater the likelihood was that the motion would be biologically relevant. Therefore, we focus on the top 9 component vectors, which cover ∼75% of total variance for each state, in greater detail.

Molecular dynamics simulations

All-atom MD simulations were prepared from the O/pO structure of IDE (PDB: 7RZK) and necessary CHARMM force field PARAM36 files using QwikMD (Ribeiro et al, 2016). Gaps in the structure were modeled in using the AlphaFold model of IDE. A three-step minimization-annealing-equilibration process was used to generate an equilibrated system. Simulations were performed under NPT (constant numbers of particles N, pressure P, and temperature T) conditions at 310 K and 1 atm with periodic boundary conditions in NAMD3.0 (Phillips et al, 2005). Explicit solvent was described with the TIP3P model (Jorgensen et al, 1983). Custom TCL scripts in VMD were used to calculate the D1-D4 centers-of-mass distance and D1-D2-D3-D4 centers-of-mass dihedral angle (Supplementary file 9). Residues 964-988 in D4 that are mostly absence in all static structures of IDE were built using Alpha-fold, which is highly flexible in our MD simulations. They were omitted from the analysis of MD simulations, as the inclusion of this flexible loop significantly altered the center-of-mass independent of global conformational change.

Initial structure of insulin bound IDE, for Upside simulations, was generated using PYMOL that started with IDE closed monomer (PDB code: 6BF8). Intact insulin (PDB code: 4INS) was fixed by mutation of residue 30 of insulin chain B back to threonine and modeled into the catalytic chamber based on the crystal structure of IDE in complex with the partially unfolded insulin (PDB code: 2WBY). The upside simulation was run with 28 replicates at TUpside=0.8 using Upside 1.0 force field parameters and our total upside simulation time is approximately 28 μsec (Faruk et al., 2022; Faruk et al., 2023; Jumper et al., 2018a, b). We estimated the time step of Upside to be ∼0.1 picosecond based on folding rates and chain motions. Custom R, and Python scripts were used for data analysis and presentation (Supplementary files 10-12).

Enzymatic assays

IDE constructs were exchanged into activity buffer comprised of 25 mM Tris-HCl pH 7.5, 150 mM NaCl, and 10 mM ZnCl2. MCA-RPPGFSAFK(Dnp) was purchased from Enzo Life Sciences (BMLP2270001), resuspended in dimethyl sulfoxide at a concentration of 5 mM and diluted to 5 μM with activity buffer. For activity measurements, 5 μM MCA-RPPGFSAFK(Dnp) was mixed with the desired IDE construct at concentrations ranging from 1-100 nM in the presence or absence of 1 mM ATP in a 200 μl reaction. Fluorescence was monitored every 30 sec for 30 minutes with excitation/emission wavelengths of 320/405 nm at 37° C. Initial velocity was calculated during the linear range. Insulin was purchased from MP Biomedicals (#0219390010), resuspended in 0.01N HCl at 1 mM and diluted to desired concentration with activity buffer. For competition assays, 5 nM IDE construct and 5 μM MCA-RPPGFSAFK(Dnp) was mixed with insulin ranging from 0-100 μM. Fluorescence was monitored every 30 sec for 30 minutes with excitation/emission wavelengths of 320/405 nm at 37° C. Initial velocity was calculated during the linear range for each concentration of insulin and normalized relative to the velocity of the respective construct in the absence of insulin to maintain consistency across plates, yielding values for relative inhibition which were plotted vs [insulin] or [Aβ] and fit to the Michaelis-Menten equation in xmgrace to generate apparent Ki values. All experiments were performed in triplicate.

Size-exclusion chromatography coupled small-angle X-ray scattering

SAXS/WAXS data collection was employed the Life Sciences X-ray Scattering (LIX) Beamline at the National Synchrotron Light Source II (NSLS II) at Brookhaven National Laboratory in Upton, NY. Briefly, 60ul of samples in solution were pipetted into PCR tubes, placed into a Bio-Inert Agilent 1260 Infinity II HPLC multisampler and measured using the isocratic SEC-SAXS format at the beamline (Yang et al, 2021). 50ul of sample was injected into the Phenomenex Biozen dSEC2 3 μm bead size with 200Å pore size column was utilized at a flow rate of 0.35mL/min for 25 minutes. SAXS and WAXS data are collected simultaneously on a Pilatus 1M (SAXS) and Pilatus 900K (WAXS) detectors with a 2 second exposure (Yang et al, 2020). Data from both detectors is then scaled and merged. Intensity is normalized using the water peak height at 2.0 Å -1. Data processing and analysis was performed using py4xs and lixtools in jupyter lab. Buffer frames 100-125 were averaged and used for subtraction of averaged frames under the peak of interest.